---
title: "Occupancy five species. Parita Bay, Panama"
subtitle: "spatial prediction"
date: "`r Sys.Date()`"
author: 
  - name: Diego J. Lizcano
    orcid: https://orcid.org/0000-0002-9648-0576
  - name: Jorge Velásquez-Tibata
    orcid: https://orcid.org/0000-0002-7773-7348
license: CC BY-SA
toc: true
format: 
  html:
    theme: cosmo
    code-fold: true
    code-block-bg: "gray70"
---


# Read spatial layers

## grouping by 3 days

```{r}
#| warning: false
#| message: false

# load function
source("C:/CodigoR/AudubonPanama/R/matrix_creator.R")
############### load Packages
library(readr) # readr 
library(readxl) # read excel
library(tidyverse) # 
library(unmarked) # occupancy by maximun likelihood
library(sf) # simple feature
library(mapview) # view map
library(ubms) # occupancy by Bayesian made easy
library(stocc) # spatial data for example
library(terra) 
library(raster)
library(RColorBrewer)
library(rasterVis)


############### load data
# old 
# parita_data_full <- read_delim("C:/CodigoR/AudubonPanama/data/all_records_audubon_panama_to_occu.csv", 
#                               delim = "\t", escape_double = FALSE, 
#                               col_types = cols(eventDate = col_date(format = "%Y-%m-%d")), 
#                               trim_ws = TRUE)

# new
parita_data_full <- read_excel("C:/CodigoR/AudubonPanama/data/all_records_audubon_panama_Darwin_Core_revision.xlsx", 
    col_types = c("text", "text", "text", 
        "date", "text", "text", "numeric", 
        "numeric", "text", "text", "numeric", 
        "text", "numeric", "numeric", "text", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "text", "text", 
        "numeric", "text", "numeric", "text", 
        "text", "text", "text", "text", "numeric", 
        "text", "numeric", "text", "numeric", 
        "text", "text"))

covs <- read_csv("C:/CodigoR/AudubonPanama/shp/sites_covs_parita_nona.csv")

## Nyctidromus albicollis
## Aramus guarauna
## Buteogallus anthracinus
## Dryocopus lineatus
## Piranga rubra

# filter by threshold and species
parita_data_5sp <-  parita_data_full %>% 
                      filter(confidence >= threshold) %>%
                      filter(scientificName == c("Nyctidromus albicollis", 
                                                 "Aramus guarauna",
                                                 "Buteogallus anthracinus",
                                                 "Dryocopus lineatus",
                                                 "Piranga rubra"
                                                 ))  

# insert Max and min to get matrix limits
# start_date
parita_data2 <- parita_data_5sp %>% 
  group_by(locationID) %>% 
  mutate(start_date = min(eventDate),
         end_date= max(eventDate))

# apply function to get matrix per species
parita <- f.matrix.creator2(data = parita_data2, year = 2023)


# centroide for terra
# centoide <- centroids(puntos, TRUE)
centroide <- c(mean(as.numeric(parita_data_5sp$decimalLongitude)), mean(as.numeric(parita_data_5sp$decimalLatitude)))
clip_window <- extent(-80.67105, -80.05822, 7.713744, 8.489888) # 7.932311, -80.612233  8.360073, -80.180109
bb <- c(-80.612233, -80.180109, 7.932311, 8.360073)
srtm <- raster::getData('SRTM', lon=centroide[1], lat=centroide[2])

# crop the  raster using the vector extent
srtm_crop <- raster::crop(srtm, clip_window)
 #rast(srtm_crop) # convert to terra


# altitud <- elevation_3s(-72.893262, 7.664081007, path="data")
# altitud <- as.numeric(Camptostoma_obsoletum_occu_dia$Altitude)

# convierte covs a puntos terra
puntos <- vect(covs, geom=c("Longitude", "Latitude"), crs="EPSG:4326")
# convierte a sf
puntos_sf <- sf::st_as_sf(puntos)
# extract elev from raster
cam_covs <- raster::extract(srtm_crop, puntos_sf)

BGB_Spawn <- rast("C:/CodigoR/AudubonPanama/raster/BGB Spawn.tif")
AGB_Spawn <- rast("C:/CodigoR/AudubonPanama/raster/AGB Spawn.tif")
NDVI <- rast("C:/CodigoR/AudubonPanama/raster/S2_NDVI_median_v2.tif")
roads <- rast("C:/CodigoR/AudubonPanama/raster/roads_final_v2.tif")
carbon_stock <- rast("C:/CodigoR/AudubonPanama/raster/soil_organic_carbon_stock_0-30m.tif")
canopy <- rast("C:/CodigoR/AudubonPanama/raster/canopy_height_jetz2.tif")
human_foot <- rast("C:/CodigoR/AudubonPanama/raster/human_footprint.tif")
river <- rast("C:/CodigoR/AudubonPanama/raster/rivers_final_v2.tif")

# make elevation equal
# srtm_projected <- projectRaster(srtm_crop, crs=projection(NDVI), method="ngb") 
# elev <- mask(resample(rast(srtm_projected), NDVI), NDVI)



# list of terras SpatRasters  
many_rasters <- list(BGB_Spawn, human_foot, NDVI, river, carbon_stock, canopy)
# terra stack
covs_raster <- rast(many_rasters)
names(covs_raster) <- c("BGB_Spawn", "human_foot",  "NDVI", "river", "carbon_stock",  "canopy")

# change NA to 0 
covs_raster0 <- subst(covs_raster, NA, 0)
# extract covariates
covs_raster_val <- extract(covs_raster0, puntos_sf)

plot(covs_raster)

##########
# chk names of species
# dete es hora, occur es presencia en el dia
# names(parita$dete[]) 


```


# *Buteogallus anthracinus*

### One species, single season

```{r}
#| warning: false
#| message: false

# Make UMF object
umf_y_full_2<- unmarkedFrameOccu(y=f.shrink.matrix.to9( 
                                   as.data.frame(parita$occur[[2]])
                                   )) #"Aramus guarauna"
siteCovs(umf_y_full_2) <- data.frame(BGB_Spawn = covs_raster_val$BGB_Spawn,
                                   human_foot = covs_raster_val$human_foot,
                                   NDVI = covs_raster_val$NDVI,
                                   river = covs_raster_val$river,
                                   carbon = covs_raster_val$carbon_stock,
                                   canopy = covs_raster_val$canopy
) # data.frame(Elev=full_covs$Elev) # Full


plot(umf_y_full_2, main="Buteogallus anthracinus")  #""Buteogallus anthracinus 

# obscanop<-matrix(rep(covs_raster_val$canopy_height_jetz2,27),ncol = 27) # make elevetion per observation

# detection
obsCovs(umf_y_full_2) <- list(hour = data.frame(f.shrink.matrix.h.to9(parita$dete[[2]]), 16, 9),
                              canopy = covs_raster_val$canopy ) # hora


# fit unmarked
fit_unm_sp2_0 <- unmarked::occu(~1~1, data=umf_y_full_2) # ok!
#fit_unm_sp2_0h <- unmarked::occu(~scale(hour)~1, data=umf_y_full_2) # It work!
# fit_unm_sp2_0c <- unmarked::occu(~scale(canopy)~1, data=umf_y_full_2) # It work!
fit_unm_sp2_1 <- unmarked::occu(~1~scale(BGB_Spawn), data=umf_y_full_2)
fit_unm_sp2_2 <- unmarked::occu(~1~scale(human_foot), data=umf_y_full_2)
fit_unm_sp2_3 <- unmarked::occu(~1~scale(NDVI), data=umf_y_full_2) # ok!
fit_unm_sp2_4 <- unmarked::occu(~1~scale(river), data=umf_y_full_2)
fit_unm_sp2_4s <- unmarked::occu(~1~scale(I(river^2)), data=umf_y_full_2)
fit_unm_sp2_5 <- unmarked::occu(~1~scale(carbon), data=umf_y_full_2)
fit_unm_sp2_6 <- unmarked::occu(~1~scale(canopy), data=umf_y_full_2)
fit_unm_sp2_7 <- unmarked::occu(~1~scale(canopy)+
                                scale(BGB_Spawn), data=umf_y_full_2)
fit_unm_sp2_8 <- unmarked::occu(~1~scale(BGB_Spawn)+
                                  scale(human_foot), data=umf_y_full_2, 
                                starts=c(1,1,1,1), method = "BFGS",
                                control = list(maxit = 20000),
                                engine = "C")
fit_unm_sp2_9 <- unmarked::occu(~1~scale(canopy)+
                                  scale(BGB_Spawn)+
                                  scale(carbon)+
                                  scale(river)+
                                  scale(NDVI)+
                                  scale(human_foot), data=umf_y_full_2)


# # fit list for detection
# fms1<-fitList("p(.) Ocu(.)"=fit_unm_sp2_0,
#               #"p(h) Ocu(.)"=fit_unm_sp2_0h, # NA problem!
#               #"p(c) Ocu(.)"=fit_unm_sp2_0c)#,
#               "p(.) Ocu(BGB_Spawn)"=fit_unm_sp2_1,
#               "p(.) Ocu(human_foot)"=fit_unm_sp2_2,
#               "p(.) Ocu(NDVI)"=fit_unm_sp2_3,
#               "p(.) Ocu(river)"=fit_unm_sp2_4,
#               "p() Ocu(carbon)"=fit_unm_sp2_5,
#               "p() Ocu(canopy)"=fit_unm_sp2_6,
#               )

# model selection detection unmarked
# modSel(fms1) #problem!.. null model ranks better

# fit list for occupancy
fms2<-fitList("p(.) Ocu(.)"=fit_unm_sp2_0,
              #"p(h) Ocu(.)"=fit_unm_sp2_0h, #,
              #"p(c) Ocu(.)"=fit_unm_sp2_0c,
              "p(.) Ocu(BGB_Spawn)"=fit_unm_sp2_1,
              "p(.) Ocu(human_foot)"=fit_unm_sp2_2,
              "p(.) Ocu(NDVI)"=fit_unm_sp2_3,
              "p(.) Ocu(river)"=fit_unm_sp2_4,
              #"p(.) Ocu(river^2)"=fit_unm_sp2_4s,
              "p() Ocu(carbon)"=fit_unm_sp2_5,
              "p() Ocu(canopy)"=fit_unm_sp2_6,
              "p() Ocu(canopy+BGB_Spawn)"=fit_unm_sp2_7,
              "p() Ocu(BGB_Spawn+human_foot)"=fit_unm_sp2_8,
              "p() Ocu(ALL)"=fit_unm_sp2_9
              )

# model selection detection unmarked
modSel(fms2)

summary(fit_unm_sp2_0) #### Null MODEL !!!!!!!!!!

print("Predicted Occupancy values")
predict(fit_unm_sp2_0, type="state")

# ######## ubms, bayesian occupancy
# # Double formula: first part is for detection, second for occupancy
# #  form <- ~detCov1 + detCov2 ~habCov1 + habCov2 + RSR(x, y, threshold=1)
# 
# # fit stan
# fit_stan_sp2_0 <- stan_occu(~1~1, data=umf_y_full_2, chains=3, iter=1000, cores=3)
# fit_stan_sp2_0h <- stan_occu(~scale(hour)~1, data=umf_y_full_2, chains=3, iter=1000, cores=3)
# fit_stan_sp2_0c <- stan_occu(~scale(canopy)~1, data=umf_y_full_2, chains=3, iter=1000, cores=3)
# fit_stan_sp2_1 <- stan_occu(~scale(hour)~scale(elevation), data=umf_y_full_2, chains=3, iter=1000, cores=3)
# fit_stan_sp2_2 <- stan_occu(~scale(hour)~scale(AGB_Spawn), data=umf_y_full_2, chains=3, iter=1000, cores=3)
# fit_stan_sp2_3 <- stan_occu(~scale(hour)~scale(roads), data=umf_y_full_2, chains=3, iter=1000, cores=3)
# fit_stan_sp2_4 <- stan_occu(~scale(hour)~scale(carbon), data=umf_y_full_2, chains=3, iter=1000, cores=3)
# fit_stan_sp2_5 <- stan_occu(~scale(hour)~scale(NDVI), data=umf_y_full_2, chains=3, iter=1000, cores=3)
# fit_stan_sp2_6 <- stan_occu(~scale(hour)~scale(canopy), data=umf_y_full_2, chains=3, iter=1000, cores=3)
# 
# 
# mods2 <- fitList(fit_stan_sp2_0,
#                 fit_stan_sp2_0h,
#                 fit_stan_sp2_0c,
#                 fit_stan_sp2_1,
#                 fit_stan_sp2_2,
#                 fit_stan_sp2_3,
#                 fit_stan_sp2_4,
#                 fit_stan_sp2_5,
#                 fit_stan_sp2_6)
# 
# # model selection
# round(modSel(mods2), 3)

# # plot top model
# print("Hour")
# plot_effects(fit_stan_sp2_5, "det") # Detection
# print("NDVI")
# plot_effects(fit_stan_sp2_5, "state") # Occupancy

```

### Spatial Prediction

```{r}
#| warning: false
#| message: false

#### predict

# 
# # srtm_crop_s <- mask(elev, NDVI)#stack(raster(elev))#, 
# #scale(roughness)) # scale altitud
# # names(srtm_crop_s) <- c("elevation")#, "roughness")
# # crs(srtm_crop_s) <- "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0"
# nd2 <- NDVI # mask(elev, NDVI) #data.frame(srtm_crop_s, hour=0.5)
# names(nd2) <- "NDVI"
# # newd <- data.frame(srtm_crop_s, hour=0.5)
# pred_psi_s_unm <-predict(fit_unm_sp2_5, type="state", newdata=nd2)
# #pred_psi_s_stan <-predict(fit_stan_sp2_1, submodel="state", newdata=nd, na.rm=FALSE)
# #pred_psi_ras <- raster(srtm_crop_s) # make raster = to elev 
# #pred_psi_ras[] <- pred_psi_s_stan$Predicted # drape on top
# 
# 
# # pred_psi_r <- pred_psi_s # * sd(full_covs$elevation) + mean(full_covs$elevation)
# # crs(pred_psi_ras) <- "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0"
# clr <- colorRampPalette(brewer.pal(9, "YlGn"))
# # mapview (pred_psi_r[[1]], col.regions = clr,  legend = TRUE, alpha=0.7)
# # plot(pred_psi_s[[1]], main="Occupancy")
# levelplot(pred_psi_s_unm[[1]], par.settings = YlOrRdTheme(), margin=FALSE, main="Ocupancy")
# 
# pal = mapviewPalette("mapviewSpectralColors")
# mapview(raster(pred_psi_s_unm[[1]]), map.types = c("Esri.WorldImagery"), alpha=0.5, col.regions = pal(10)) + mapview(puntos_sf)


```


# *Dryocopus lineatus*

### One species, single season


```{r}
#| warning: false
#| message: false

# Make UMF object
# Make UMF object
umf_y_full_3<- unmarkedFrameOccu(y=f.shrink.matrix.to9( 
                                   as.data.frame(parita$occur[[3]])
                                   )) #"Aramus guarauna"
siteCovs(umf_y_full_3) <- data.frame(BGB_Spawn = covs_raster_val$BGB_Spawn,
                                   human_foot = covs_raster_val$human_foot,
                                   NDVI = covs_raster_val$NDVI,
                                   river = covs_raster_val$river,
                                   carbon_stock = covs_raster_val$carbon_stock,
                                   canopy = covs_raster_val$canopy
) # data.frame(Elev=full_covs$Elev) # Full


plot(umf_y_full_3, main="Dryocopus lineatus")  #""Buteogallus anthracinus 

# obscanop<-matrix(rep(covs_raster_val$canopy_height_jetz2,27),ncol = 27) # make elevetion per observation

# detection
obsCovs(umf_y_full_3) <- list(hour = data.frame(f.shrink.matrix.h.to9(parita$dete[[3]]), 16, 9),
                              canopy = covs_raster_val$canopy ) # hora


# fit unmarked
fit_unm_sp3_0 <- unmarked::occu(~1~1, data=umf_y_full_3) # ok!
#fit_unm_sp2_0h <- unmarked::occu(~scale(hour)~1, data=umf_y_full_2) # It work!
# fit_unm_sp2_0c <- unmarked::occu(~scale(canopy)~1, data=umf_y_full_2) # It work!
fit_unm_sp3_1 <- unmarked::occu(~1~scale(BGB_Spawn), data=umf_y_full_3,  starts=c(1,1,1), method = "BFGS",
               control = list(maxit = 20000),
               engine = "C")

fit_unm_sp3_2 <- unmarked::occu(~1~scale(human_foot), data=umf_y_full_3)
fit_unm_sp3_3 <- unmarked::occu(~1~scale(NDVI), data=umf_y_full_3) # ok!
fit_unm_sp3_4 <- unmarked::occu(~1~scale(river), data=umf_y_full_3)
fit_unm_sp3_4s <- unmarked::occu(~1~scale(I(river^2)), data=umf_y_full_3)
fit_unm_sp3_5 <- unmarked::occu(~1~scale(carbon_stock), data=umf_y_full_3)
fit_unm_sp3_6 <- unmarked::occu(~1~scale(canopy), data=umf_y_full_3)
fit_unm_sp3_7 <- unmarked::occu(~1~scale(canopy)+
                                  scale(BGB_Spawn), data=umf_y_full_3)
fit_unm_sp3_8 <- unmarked::occu(~1~scale(BGB_Spawn)+
                                  scale(human_foot), data=umf_y_full_3, 
                                starts=c(1,1,1,1), method = "BFGS",
                                control = list(maxit = 20000),
                                engine = "C")
fit_unm_sp3_9 <- unmarked::occu(~1~scale(river)+ # non carbon
                                  scale(NDVI)+ 
                                  scale(human_foot), data=umf_y_full_3)
fit_unm_sp3_10 <- unmarked::occu(~1~scale(canopy)+ # all carbon
                                  scale(BGB_Spawn)+ 
                                  scale(carbon_stock), data=umf_y_full_3,
                                starts=c(1,1,1,1,1), method = "BFGS",
                                control = list(maxit = 20000),
                                engine = "C")
fit_unm_sp3_11 <- unmarked::occu(~1~scale(canopy)+
                                  scale(BGB_Spawn)+ 
                                  scale(carbon_stock)+
                                  scale(river)+ 
                                  scale(NDVI)+ 
                                  scale(human_foot), data=umf_y_full_3)

# # fit list for detection
# fms1<-fitList("p(.) Ocu(.)"=fit_unm_sp2_0,
#               #"p(h) Ocu(.)"=fit_unm_sp2_0h, # NA problem!
#               #"p(c) Ocu(.)"=fit_unm_sp2_0c)#,
#               "p(.) Ocu(BGB_Spawn)"=fit_unm_sp2_1,
#               "p(.) Ocu(human_foot)"=fit_unm_sp2_2,
#               "p(.) Ocu(NDVI)"=fit_unm_sp2_3,
#               "p(.) Ocu(river)"=fit_unm_sp2_4,
#               "p() Ocu(carbon)"=fit_unm_sp2_5,
#               "p() Ocu(canopy)"=fit_unm_sp2_6,
#               )

# model selection detection unmarked
# modSel(fms1) #problem!.. null model ranks better

# fit list for occupancy
fms3<-fitList("p(.) Ocu(.)"=fit_unm_sp3_0,
              #"p(h) Ocu(.)"=fit_unm_sp2_0h, #,
              #"p(c) Ocu(.)"=fit_unm_sp2_0c,
              "p(.) Ocu(BGB_Spawn)"=fit_unm_sp3_1,
              "p(.) Ocu(human_foot)"=fit_unm_sp3_2,
              "p(.) Ocu(NDVI)"=fit_unm_sp3_3,
              "p(.) Ocu(river)"=fit_unm_sp3_4,
              #"p(.) Ocu(river^2)"=fit_unm_sp2_4s,
              "p() Ocu(carbon)"=fit_unm_sp3_5,
              "p() Ocu(canopy)"=fit_unm_sp3_6,
              "p() Ocu(canopy+BGB_Spawn)"=fit_unm_sp3_7,
              "p() Ocu(BGB_Spawn+human_foot)"=fit_unm_sp3_8,
              "p() Ocu(river+NDVI+human_foot)"=fit_unm_sp3_9,
              "p() Ocu(canopy+BGB_Spawn+carbon)"=fit_unm_sp3_10,
              "p() Ocu(ALL)"=fit_unm_sp3_11
              )

# model selection detection unmarked
modSel(fms3)

summary(fit_unm_sp3_10)

pb_f <- parboot(fit_unm_sp3_10, nsim=500, report=10) 
plot (pb_f)


newdat_range<-data.frame(BGB_Spawn=seq(min(covs_raster_val$BGB_Spawn),                                       max(covs_raster_val$BGB_Spawn),length=100),
                         human_foot=seq(min(covs_raster_val$human_foot),                                       max(covs_raster_val$human_foot),length=100),
                         NDVI=seq(min(covs_raster_val$NDVI),                                                  max(covs_raster_val$NDVI),length=100),
                         river=seq(min(covs_raster_val$river),                                                  max(covs_raster_val$river),length=100),
                         carbon_stock=seq(min(covs_raster_val$carbon),                                                  max(covs_raster_val$carbon),length=100),
                         canopy=seq(min(covs_raster_val$canopy),                                                  max(covs_raster_val$canopy),length=100)
                         )


pred_psi <-predict(fit_unm_sp3_10, type="state", newdata=newdat_range, appendData=TRUE) 

###  Plot occupancy en escala original
plot(Predicted~BGB_Spawn, pred_psi,type="l",col="blue", ylim=c(0,0.95),
     xlab="canopy",
     ylab="Probabilidad de ocupación")
lines(lower~canopy, pred_psi,type="l",col=gray(0.5))
lines(upper~canopy, pred_psi,type="l",col=gray(0.5))



# # fit stan
# fit_stan_sp3_0 <- stan_occu(~1~1, data=umf_y_full_3, chains=3, iter=1000, cores=3)
# fit_stan_sp3_0h <- stan_occu(~scale(hour)~1, data=umf_y_full_3, chains=3, iter=1000, cores=3)
# fit_stan_sp3_1 <- stan_occu(~scale(hour)~scale(elevation), data=umf_y_full_3, chains=3, iter=1000, cores=3)
# fit_stan_sp3_2 <- stan_occu(~scale(hour)~scale(AGB_Spawn), data=umf_y_full_3, chains=3, iter=1000, cores=3)
# fit_stan_sp3_3 <- stan_occu(~scale(hour)~scale(roads), data=umf_y_full_3, chains=3, iter=1000, cores=3)
# fit_stan_sp3_4 <- stan_occu(~scale(hour)~scale(carbon), data=umf_y_full_3, chains=3, iter=1000, cores=3)
# fit_stan_sp3_5 <- stan_occu(~scale(hour)~scale(NDVI), data=umf_y_full_3, chains=3, iter=1000, cores=3)
# fit_stan_sp3_6 <- stan_occu(~scale(hour)~scale(canopy), data=umf_y_full_3, chains=3, iter=1000, cores=3)
# 
# 
# mods3 <- fitList(fit_stan_sp3_0,
#                 fit_stan_sp3_0h,
#                 fit_stan_sp3_1,
#                 fit_stan_sp3_2,
#                 fit_stan_sp3_3,
#                 fit_stan_sp3_4,
#                 fit_stan_sp3_5,
#                 fit_stan_sp3_6)
# 
# # model selection
# round(modSel(mods3), 3) #AGB_Spawn
# 
# # plot top model
# print("Hour")
# plot_effects(fit_stan_sp3_5, "det") # Detectiom
# print("NDVI")
# plot_effects(fit_stan_sp3_5, "state") # Ocupancy

```


### Spatial Prediction

```{r}
#| warning: false
#| message: false

#spatial
# occuPred <- unmarked::predict(fit_unm_sp3_10,
#                     type = "state",
#                     newdata = stack(raster(covs_raster[[1]]),
#                                     raster(covs_raster[[5]]),
#                                     raster(covs_raster[[6]])), 
#                     na.rm = TRUE,
#                     inf.rm = TRUE,
#                     progress = 'text')

newdata_terra = c(covs_raster[[1]],
                 covs_raster[[5]],
                 covs_raster[[6]])

occuPred <- terra::predict(newdata_terra, fit_unm_sp3_10, type="state", cores=3, na.rm=TRUE, progress = 'text')


#### predict

# srtm_crop_s <- mask(elev, NDVI)#stack(raster(elev))#, 
# scale(roughness)) # scale altitud
# names(srtm_crop_s) <- c("elevation")#, "roughness")
# crs(srtm_crop_s) <- "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0"
#nd3 <- NDVI #mask(AGB_Spawn, NDVI) #data.frame(srtm_crop_s, hour=0.5)
#names(nd3) <- "NDVI"
# newd <- data.frame(srtm_crop_s, hour=0.5)
#pred_psi_s_unm <-predict(fit_unm_sp3_5, type="state", newdata=nd3)
#pred_psi_s_stan <-predict(fit_stan_sp2_1, submodel="state", newdata=nd, na.rm=FALSE)
#pred_psi_ras <- raster(srtm_crop_s) # make raster = to elev 
#pred_psi_ras[] <- pred_psi_s_stan$Predicted # drape on top


# pred_psi_r <- pred_psi_s # * sd(full_covs$elevation) + mean(full_covs$elevation)
# crs(pred_psi_ras) <- "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0"
clr <- colorRampPalette(brewer.pal(9, "YlGn"))
# mapview (pred_psi_r[[1]], col.regions = clr,  legend = TRUE, alpha=0.7)
# plot(pred_psi_s[[1]], main="Occupancy")
levelplot(occuPred[[1]], par.settings = YlOrRdTheme(), margin=FALSE, main="Ocupancy")

pal = mapviewPalette("mapviewSpectralColors")
mapview(raster(occuPred[[1]]), map.types = c("Esri.WorldImagery"), alpha=0.5, col.regions = pal(10)) + mapview(puntos_sf)


```



# *Nyctidromus albicollis*

### One species, single season


```{r}
#| warning: false
#| message: false

umf_y_full_4<- unmarkedFrameOccu(y=f.shrink.matrix.to9( 
                                   as.data.frame(parita$occur[[4]])
                                   )) #"Aramus guarauna"
siteCovs(umf_y_full_4) <- data.frame(BGB_Spawn = covs_raster_val$BGB_Spawn,
                                   human_foot = covs_raster_val$human_foot,
                                   NDVI = covs_raster_val$NDVI,
                                   river = covs_raster_val$river,
                                   carbon_stock = covs_raster_val$carbon_stock,
                                   canopy = covs_raster_val$canopy
) # data.frame(Elev=full_covs$Elev) # Full


plot(umf_y_full_4, main="Nyctidromus albicollis")  #""Buteogallus anthracinus 

# obscanop<-matrix(rep(covs_raster_val$canopy_height_jetz2,27),ncol = 27) # make elevetion per observation

# detection
obsCovs(umf_y_full_4) <- list(hour = data.frame(f.shrink.matrix.h.to9(parita$dete[[4]]), 16, 9),
                              canopy = covs_raster_val$canopy ) # hora


# fit unmarked
fit_unm_sp4_0 <- unmarked::occu(~1~1, data=umf_y_full_4) # ok!
#fit_unm_sp2_0h <- unmarked::occu(~scale(hour)~1, data=umf_y_full_2) # It work!
# fit_unm_sp2_0c <- unmarked::occu(~scale(canopy)~1, data=umf_y_full_2) # It work!
fit_unm_sp4_1 <- unmarked::occu(~1~scale(BGB_Spawn), data=umf_y_full_4,  starts=c(1,1,1), method = "BFGS",
               control = list(maxit = 20000),
               engine = "C")

fit_unm_sp4_2 <- unmarked::occu(~1~scale(human_foot), data=umf_y_full_4)
fit_unm_sp4_3 <- unmarked::occu(~1~scale(NDVI), data=umf_y_full_4) # ok!
fit_unm_sp4_4 <- unmarked::occu(~1~scale(river), data=umf_y_full_4)
fit_unm_sp4_4s <- unmarked::occu(~1~scale(I(river^2)), data=umf_y_full_4)
fit_unm_sp4_5 <- unmarked::occu(~1~scale(carbon_stock), data=umf_y_full_4)
fit_unm_sp4_6 <- unmarked::occu(~1~scale(canopy), data=umf_y_full_4)
fit_unm_sp4_7 <- unmarked::occu(~1~scale(canopy)+
                                  scale(BGB_Spawn), data=umf_y_full_4)
fit_unm_sp4_8 <- unmarked::occu(~1~scale(BGB_Spawn)+
                                  scale(human_foot), data=umf_y_full_4, 
                                starts=c(1,1,1,1), method = "BFGS",
                                control = list(maxit = 20000),
                                engine = "C")
fit_unm_sp4_9 <- unmarked::occu(~1~scale(river)+ # non carbon
                                  scale(NDVI)+ 
                                  scale(human_foot), data=umf_y_full_4)
fit_unm_sp4_10 <- unmarked::occu(~1~scale(canopy)+ # all carbon
                                  scale(BGB_Spawn)+ 
                                  scale(carbon_stock), data=umf_y_full_4,
                                starts=c(1,1,1,1,1), method = "BFGS",
                                control = list(maxit = 20000),
                                engine = "C")
fit_unm_sp4_11 <- unmarked::occu(~1~scale(canopy)+
                                  scale(BGB_Spawn)+ 
                                  scale(carbon_stock)+
                                  scale(river)+ 
                                  scale(NDVI)+ 
                                  scale(human_foot), data=umf_y_full_4)


# # fit list for detection
# fms1<-fitList("p(.) Ocu(.)"=fit_unm_sp2_0,
#               #"p(h) Ocu(.)"=fit_unm_sp2_0h, # NA problem!
#               #"p(c) Ocu(.)"=fit_unm_sp2_0c)#,
#               "p(.) Ocu(BGB_Spawn)"=fit_unm_sp2_1,
#               "p(.) Ocu(human_foot)"=fit_unm_sp2_2,
#               "p(.) Ocu(NDVI)"=fit_unm_sp2_3,
#               "p(.) Ocu(river)"=fit_unm_sp2_4,
#               "p() Ocu(carbon)"=fit_unm_sp2_5,
#               "p() Ocu(canopy)"=fit_unm_sp2_6,
#               )

# model selection detection unmarked
# modSel(fms1) #problem!.. null model ranks better

# fit list for occupancy
fms4<-fitList("p(.) Ocu(.)"=fit_unm_sp4_0,
              #"p(h) Ocu(.)"=fit_unm_sp2_0h, #,
              #"p(c) Ocu(.)"=fit_unm_sp2_0c,
              "p(.) Ocu(BGB_Spawn)"=fit_unm_sp4_1,
              "p(.) Ocu(human_foot)"=fit_unm_sp4_2,
              "p(.) Ocu(NDVI)"=fit_unm_sp4_3,
              "p(.) Ocu(river)"=fit_unm_sp4_4,
              #"p(.) Ocu(river^2)"=fit_unm_sp4_4s,
              "p() Ocu(carbon)"=fit_unm_sp4_5,
              "p() Ocu(canopy)"=fit_unm_sp4_6,
              "p() Ocu(canopy+BGB_Spawn)"=fit_unm_sp4_7,
              "p() Ocu(BGB_Spawn+human_foot)"=fit_unm_sp4_8,
              "p() Ocu(river+NDVI+human_foot)"=fit_unm_sp4_9,
              "p() Ocu(canopy+BGB_Spawn+carbon)"=fit_unm_sp4_10,
              "p() Ocu(ALL)"=fit_unm_sp4_11
              )

# model selection detection unmarked
modSel(fms4)

summary(fit_unm_sp4_4)

#pb_f <- parboot(fit_unm_sp4_4, nsim=500, report=10) 
#plot (pb_f)

############## function to evaluate, model
fitstats <- function(model.name, 
                     method = "nonparboot") {
  observed <- getY(model.name@data)
  expected <- fitted(model.name)
  resids <- residuals(model.name,
                      method = "nonparboot")
  sse <- sum(resids^2,
             na.rm = TRUE)
  chisq <- sum((observed - expected)^2 / expected,
               na.rm = TRUE)
  freeTuke <- sum((sqrt(observed) - sqrt(expected))^2, 
                  na.rm = TRUE)
  out <- c(SSE = sse,
           Chisq = chisq,
           freemanTukey = freeTuke)
  return(out)
}

# wait!...
pb <- parboot(fit_unm_sp4_4,
              fitstats,
              nsim = 1000,
              report = TRUE,
              method = "nonparboot")
pb

par(mfrow = c(3,1)) # 3 en una
plot(pb,
     main = "All",
     xlab = c("SSE", "Chisq", "FT"))
dev.off() # reset par

############# histogram evaluation
pb_f <- parboot(fit_unm_sp4_4, nsim=500, report=10) 
plot (pb_f)



newdat_range<-data.frame(BGB_Spawn=seq(min(covs_raster_val$BGB_Spawn),                                       max(covs_raster_val$BGB_Spawn),length=100),
                         human_foot=seq(min(covs_raster_val$human_foot),                                       max(covs_raster_val$human_foot),length=100),
                         NDVI=seq(min(covs_raster_val$NDVI),                                                  max(covs_raster_val$NDVI),length=100),
                         river=seq(min(covs_raster_val$river),                                                  max(covs_raster_val$river),length=100),
                         carbon_stock=seq(min(covs_raster_val$carbon_stock),                                                  max(covs_raster_val$carbon),length=100),
                         canopy=seq(min(covs_raster_val$canopy),                                                  max(covs_raster_val$canopy),length=100)
                         )

pred_psi <-predict(fit_unm_sp4_4, type="state", newdata=newdat_range, appendData=TRUE) 

###  Plot occupancy en escala original
plot(Predicted~river, pred_psi,type="l",col="blue", ylim=c(0,0.95),
     xlab="river",
     ylab="Probabilidad de ocupación")
lines(lower~river, pred_psi,type="l",col=gray(0.5))
lines(upper~river, pred_psi,type="l",col=gray(0.5))

# # fit stan
# fit_stan_sp3_0 <- stan_occu(~1~1, data=umf_y_full_3, chains=3, iter=1000, cores=3)
# fit_stan_sp3_0h <- stan_occu(~scale(hour)~1, data=umf_y_full_3, chains=3, iter=1000, cores=3)
# fit_stan_sp3_1 <- stan_occu(~scale(hour)~scale(elevation), data=umf_y_full_3, chains=3, iter=1000, cores=3)
# fit_stan_sp3_2 <- stan_occu(~scale(hour)~scale(AGB_Spawn), data=umf_y_full_3, chains=3, iter=1000, cores=3)
# fit_stan_sp3_3 <- stan_occu(~scale(hour)~scale(roads), data=umf_y_full_3, chains=3, iter=1000, cores=3)
# fit_stan_sp3_4 <- stan_occu(~scale(hour)~scale(carbon), data=umf_y_full_3, chains=3, iter=1000, cores=3)
# fit_stan_sp3_5 <- stan_occu(~scale(hour)~scale(NDVI), data=umf_y_full_3, chains=3, iter=1000, cores=3)
# fit_stan_sp3_6 <- stan_occu(~scale(hour)~scale(canopy), data=umf_y_full_3, chains=3, iter=1000, cores=3)
# 
# 
# mods3 <- fitList(fit_stan_sp3_0,
#                 fit_stan_sp3_0h,
#                 fit_stan_sp3_1,
#                 fit_stan_sp3_2,
#                 fit_stan_sp3_3,
#                 fit_stan_sp3_4,
#                 fit_stan_sp3_5,
#                 fit_stan_sp3_6)
# 
# # model selection
# round(modSel(mods3), 3) #AGB_Spawn
# 
# # plot top model
# print("Hour")
# plot_effects(fit_stan_sp3_5, "det") # Detectiom
# print("NDVI")
# plot_effects(fit_stan_sp3_5, "state") # Ocupancy

```


### Spatial Prediction

```{r}
#| warning: false
#| message: false

#### predict

# srtm_crop_s <- mask(elev, NDVI)#stack(raster(elev))#, 
# scale(roughness)) # scale altitud
# names(srtm_crop_s) <- c("elevation")#, "roughness")
# crs(srtm_crop_s) <- "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0"
# nd4 <- mask(AGB_Spawn, NDVI) #data.frame(srtm_crop_s, hour=0.5)
# names(nd4) <- "AGB_Spawn"
# newd <- data.frame(srtm_crop_s, hour=0.5)
# pred_psi_s_unm <-predict(fit_unm_sp4_4, type="state", newdata=nd4)


# occuPred <- unmarked::predict(fit_unm_sp4_4,
#                     type = "state",
#                     newdata = covs_raster[[4]], 
#                     na.rm = TRUE,
#                     inf.rm = TRUE)

newdata_terra = c(covs_raster[[4]])#,
                 #covs_raster[[5]],
                 #covs_raster[[6]])

occuPred <- terra::predict(newdata_terra, fit_unm_sp4_4, type="state", cores=3, na.rm=TRUE, progress = 'text')



#pred_psi_s_stan <-predict(fit_stan_sp2_1, submodel="state", newdata=nd, na.rm=FALSE)
#pred_psi_ras <- raster(srtm_crop_s) # make raster = to elev 
#pred_psi_ras[] <- pred_psi_s_stan$Predicted # drape on top


# pred_psi_r <- pred_psi_s # * sd(full_covs$elevation) + mean(full_covs$elevation)
# crs(pred_psi_ras) <- "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0"
clr <- colorRampPalette(brewer.pal(9, "YlGn"))
# mapview (pred_psi_r[[1]], col.regions = clr,  legend = TRUE, alpha=0.7)
# plot(pred_psi_s[[1]], main="Occupancy")
levelplot(occuPred[[1]], par.settings = YlOrRdTheme(), margin=FALSE, main="Ocupancy")

pal = mapviewPalette("mapviewSpectralColors")
mapview(raster(occuPred[[1]]), map.types = c("Esri.WorldImagery"), alpha=0.5, col.regions = pal(10)) + mapview(puntos_sf)



```




# *Piranga rubra*

### One species, single season


```{r}
#| warning: false
#| message: false

# Make UMF object
umf_y_full_5<- unmarkedFrameOccu(y= as.data.frame(parita$occur[[5]])) #Dryocopus lineatus

siteCovs(umf_y_full_5) <- data.frame(BGB_Spawn = covs_raster_val$BGB_Spawn,
                                   human_foot = covs_raster_val$human_foot,
                                   NDVI = covs_raster_val$NDVI,
                                   river = covs_raster_val$river,
                                   carbon_stock = covs_raster_val$carbon_stock,
                                   canopy = covs_raster_val$canopy
                                   )


plot(umf_y_full_5, main="Piranga rubra")  #""Buteogallus anthracinus no converge

# detection
obsCovs(umf_y_full_5) <- list(hour = matrix(parita$dete[[5]]),
                              canopy = covs_raster_val$canopy ) # hora

# Double formula: first part is for detection, second for occupancy
#  form <- ~detCov1 + detCov2 ~habCov1 + habCov2 + RSR(x, y, threshold=1)

# fit unmarked
# fit unmarked
fit_unm_sp5_0 <- unmarked::occu(~1~1, data=umf_y_full_5) # ok!
#fit_unm_sp2_0h <- unmarked::occu(~scale(hour)~1, data=umf_y_full_2) # It work!
# fit_unm_sp2_0c <- unmarked::occu(~scale(canopy)~1, data=umf_y_full_2) # It work!
fit_unm_sp5_1 <- unmarked::occu(~1~scale(BGB_Spawn), data=umf_y_full_5,  starts=c(1,1,1), method = "BFGS",
               control = list(maxit = 20000),
               engine = "C")

fit_unm_sp5_2 <- unmarked::occu(~1~scale(human_foot), data=umf_y_full_5)
fit_unm_sp5_3 <- unmarked::occu(~1~scale(NDVI), data=umf_y_full_5) # ok!
fit_unm_sp5_4 <- unmarked::occu(~1~scale(river), data=umf_y_full_5)
fit_unm_sp5_4s <- unmarked::occu(~1~scale(I(river^2)), data=umf_y_full_5)
fit_unm_sp5_5 <- unmarked::occu(~1~scale(carbon_stock), data=umf_y_full_5)
fit_unm_sp5_6 <- unmarked::occu(~1~scale(canopy), data=umf_y_full_5)
fit_unm_sp5_7 <- unmarked::occu(~1~scale(canopy)+
                                  scale(BGB_Spawn), data=umf_y_full_5)
fit_unm_sp5_8 <- unmarked::occu(~1~scale(BGB_Spawn)+
                                  scale(human_foot), data=umf_y_full_5, 
                                starts=c(1,1,1,1), method = "BFGS",
                                control = list(maxit = 20000),
                                engine = "C")
fit_unm_sp5_9 <- unmarked::occu(~1~scale(river)+ # non carbon
                                  scale(NDVI)+ 
                                  scale(human_foot), data=umf_y_full_5)
fit_unm_sp5_10 <- unmarked::occu(~1~scale(canopy)+ # all carbon
                                  scale(BGB_Spawn)+ 
                                  scale(carbon_stock), data=umf_y_full_5,
                                starts=c(1,1,1,1,1), method = "BFGS",
                                control = list(maxit = 20000),
                                engine = "C")
fit_unm_sp5_11 <- unmarked::occu(~1~scale(canopy)+
                                  scale(BGB_Spawn)+ 
                                  scale(carbon_stock)+
                                  scale(river)+ 
                                  scale(NDVI)+ 
                                  scale(human_foot), data=umf_y_full_5)


# # fit list for detection
# fms1<-fitList("p(.) Ocu(.)"=fit_unm_sp2_0,
#               #"p(h) Ocu(.)"=fit_unm_sp2_0h, # NA problem!
#               #"p(c) Ocu(.)"=fit_unm_sp2_0c)#,
#               "p(.) Ocu(BGB_Spawn)"=fit_unm_sp2_1,
#               "p(.) Ocu(human_foot)"=fit_unm_sp2_2,
#               "p(.) Ocu(NDVI)"=fit_unm_sp2_3,
#               "p(.) Ocu(river)"=fit_unm_sp2_4,
#               "p() Ocu(carbon)"=fit_unm_sp2_5,
#               "p() Ocu(canopy)"=fit_unm_sp2_6,
#               )

# model selection detection unmarked
# modSel(fms1) #problem!.. null model ranks better

# fit list for occupancy
fms5<-fitList("p(.) Ocu(.)"=fit_unm_sp5_0,
              #"p(h) Ocu(.)"=fit_unm_sp2_0h, #,
              #"p(c) Ocu(.)"=fit_unm_sp2_0c,
              "p(.) Ocu(BGB_Spawn)"=fit_unm_sp5_1,
              "p(.) Ocu(human_foot)"=fit_unm_sp5_2,
              "p(.) Ocu(NDVI)"=fit_unm_sp5_3,
              "p(.) Ocu(river)"=fit_unm_sp5_4,
              #"p(.) Ocu(river^2)"=fit_unm_sp4_4s,
              "p() Ocu(carbon)"=fit_unm_sp5_5,
              "p() Ocu(canopy)"=fit_unm_sp5_6,
              "p() Ocu(canopy+BGB_Spawn)"=fit_unm_sp5_7,
              "p() Ocu(BGB_Spawn+human_foot)"=fit_unm_sp5_8,
              "p() Ocu(river+NDVI+human_foot)"=fit_unm_sp5_9,
              "p() Ocu(canopy+BGB_Spawn+carbon)"=fit_unm_sp5_10,
              "p() Ocu(ALL)"=fit_unm_sp5_11
              )

# model selection detection unmarked
modSel(fms5)

summary(fit_unm_sp5_7)

############# histogram evaluation
pb_f <- parboot(fit_unm_sp5_7, nsim=500, report=10) 
plot (pb_f)


pred_psi <-predict(fit_unm_sp5_7, type="state", newdata=newdat_range, appendData=TRUE) 

###  Plot occupancy en escala original
plot(Predicted~canopy, pred_psi,type="l",col="blue", ylim=c(0,0.95),
     xlab="canopy",
     ylab="Probabilidad de ocupación")
lines(lower~canopy, pred_psi,type="l",col=gray(0.5))
lines(upper~canopy, pred_psi,type="l",col=gray(0.5))

###  Plot occupancy en escala original
plot(Predicted~BGB_Spawn, pred_psi,type="l",col="blue", ylim=c(0,0.95),
     xlab="BGB_Spawn",
     ylab="Probabilidad de ocupación")
lines(lower~BGB_Spawn, pred_psi,type="l",col=gray(0.5))
lines(upper~BGB_Spawn, pred_psi,type="l",col=gray(0.5))



# 
# # fit stan
# fit_stan_sp5_0 <- stan_occu(~1~1, data=umf_y_full_5, chains=3, iter=1000, cores=3)
# fit_stan_sp5_0h <- stan_occu(~scale(hour)~1, data=umf_y_full_5, chains=3, iter=1000, cores=3)
# fit_stan_sp5_0c <- stan_occu(~scale(canopy)~1, data=umf_y_full_5, chains=3, iter=1000, cores=3)
# fit_stan_sp5_1 <- stan_occu(~scale(hour)~scale(elevation), data=umf_y_full_5, chains=3, iter=1000, cores=3)
# fit_stan_sp5_2 <- stan_occu(~scale(hour)~scale(AGB_Spawn), data=umf_y_full_5, chains=3, iter=1000, cores=3)
# fit_stan_sp5_3 <- stan_occu(~scale(hour)~scale(roads), data=umf_y_full_5, chains=3, iter=1000, cores=3)
# fit_stan_sp5_4 <- stan_occu(~scale(hour)~scale(carbon), data=umf_y_full_5, chains=3, iter=1000, cores=3)
# fit_stan_sp5_5 <- stan_occu(~scale(hour)~scale(NDVI), data=umf_y_full_5, chains=3, iter=1000, cores=3)
# fit_stan_sp5_6 <- stan_occu(~scale(hour)~scale(canopy), data=umf_y_full_5, chains=3, iter=1000, cores=3)
# 
# 
# mods5 <- fitList(fit_stan_sp5_0,
#                 fit_stan_sp5_0h,
#                 fit_stan_sp5_0c,
#                 fit_stan_sp5_1,
#                 fit_stan_sp5_2,
#                 fit_stan_sp5_3,
#                 fit_stan_sp5_4,
#                 fit_stan_sp5_5)
# 
# # model selection
# round(modSel(mods5), 3)
# 
# # plot top model
# print("Hour")
# plot_effects(fit_stan_sp5_2, "det") # Detection
# print("AGB_Spawn")
# plot_effects(fit_stan_sp5_2, "state") # Occupancy

```


### Spatial Prediction

```{r}
#| warning: false
#| message: false

#### predict

# srtm_crop_s <- mask(elev, NDVI)#stack(raster(elev))#, 
# scale(roughness)) # scale altitud
# names(srtm_crop_s) <- c("elevation")#, "roughness")
# crs(srtm_crop_s) <- "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0"
# nd5 <- mask(AGB_Spawn, NDVI) #data.frame(srtm_crop_s, hour=0.5)
# names(nd5) <- "AGB_Spawn"
# newd <- data.frame(srtm_crop_s, hour=0.5)
#pred_psi_s_unm <-predict(fit_unm_sp5_2b, type="state", newdata=nd5)
#pred_psi_s_stan <-predict(fit_stan_sp2_1, submodel="state", newdata=nd, na.rm=FALSE)
#pred_psi_ras <- raster(srtm_crop_s) # make raster = to elev 
#pred_psi_ras[] <- pred_psi_s_stan$Predicted # drape on top

newdata_terra = c(covs_raster[[1]],#,
                 covs_raster[[6]])#,
                 #covs_raster[[6]])

occuPred <- terra::predict(newdata_terra, fit_unm_sp5_7, type="state", cores=3, na.rm=TRUE, progress = 'text')



# pred_psi_r <- pred_psi_s # * sd(full_covs$elevation) + mean(full_covs$elevation)
# crs(pred_psi_ras) <- "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0"
clr <- colorRampPalette(brewer.pal(9, "YlGn"))
# mapview (pred_psi_r[[1]], col.regions = clr,  legend = TRUE, alpha=0.7)
# plot(pred_psi_s[[1]], main="Occupancy")
levelplot(occuPred[[1]], par.settings = YlOrRdTheme(), margin=FALSE, main="Ocupancy")

pal = mapviewPalette("mapviewSpectralColors")
mapview(raster(occuPred[[1]]), map.types = c("Esri.WorldImagery"), alpha=0.5, col.regions = pal(10)) + mapview(puntos_sf)


```






# *Aramus guarauna*

### One species, single season


```{r}
#| warning: false
#| message: false

# Make UMF object
umf_y_full_1<- unmarkedFrameOccu(y= as.data.frame(parita$occur[[1]])) #Aramus guarauna

siteCovs(umf_y_full_1) <- data.frame(BGB_Spawn = covs_raster_val$BGB_Spawn,
                                   human_foot = covs_raster_val$human_foot,
                                   NDVI = covs_raster_val$NDVI,
                                   river = covs_raster_val$river,
                                   carbon_stock = covs_raster_val$carbon_stock,
                                   canopy = covs_raster_val$canopy
                                   ) # data.frame(Elev=full_covs$Elev) # Full


plot(umf_y_full_1, main="Aramus guarauna")  #""Buteogallus anthracinus no converge

# detection
obsCovs(umf_y_full_1) <- list(hour = matrix(parita$dete[[1]]),
                              canopy = covs_raster_val$canopy ) # hora

# Double formula: first part is for detection, second for occupancy
#  form <- ~detCov1 + detCov2 ~habCov1 + habCov2 + RSR(x, y, threshold=1)

# fit unmarked
# fit unmarked
fit_unm_sp1_0 <- unmarked::occu(~1~1, data=umf_y_full_1) # ok!
#fit_unm_sp2_0h <- unmarked::occu(~scale(hour)~1, data=umf_y_full_2) # It work!
# fit_unm_sp2_0c <- unmarked::occu(~scale(canopy)~1, data=umf_y_full_2) # It work!
fit_unm_sp1_1 <- unmarked::occu(~1~scale(BGB_Spawn), data=umf_y_full_1,  starts=c(1,1,1), method = "BFGS",
               control = list(maxit = 20000),
               engine = "C")

fit_unm_sp1_2 <- unmarked::occu(~1~scale(human_foot), data=umf_y_full_1)
fit_unm_sp1_3 <- unmarked::occu(~1~scale(NDVI), data=umf_y_full_1) # ok!
fit_unm_sp1_4 <- unmarked::occu(~1~scale(river), data=umf_y_full_1)
fit_unm_sp1_4s <- unmarked::occu(~1~scale(I(river^2)), data=umf_y_full_1)
fit_unm_sp1_5 <- unmarked::occu(~1~scale(carbon_stock), data=umf_y_full_1)
fit_unm_sp1_6 <- unmarked::occu(~1~scale(canopy), data=umf_y_full_1)
fit_unm_sp1_7 <- unmarked::occu(~1~scale(canopy)+
                                  scale(BGB_Spawn), data=umf_y_full_1)
fit_unm_sp1_8 <- unmarked::occu(~1~scale(BGB_Spawn)+
                                  scale(human_foot), data=umf_y_full_1, 
                                starts=c(1,1,1,1), method = "BFGS",
                                control = list(maxit = 20000),
                                engine = "C")
fit_unm_sp1_9 <- unmarked::occu(~1~scale(river)+ # non carbon
                                  scale(NDVI)+ 
                                  scale(human_foot), data=umf_y_full_1)
fit_unm_sp1_10 <- unmarked::occu(~1~scale(canopy)+ # all carbon
                                  scale(BGB_Spawn)+ 
                                  scale(carbon_stock), data=umf_y_full_1,
                                starts=c(1,1,1,1,1), method = "BFGS",
                                control = list(maxit = 20000),
                                engine = "C")
fit_unm_sp1_11 <- unmarked::occu(~1~scale(canopy)+
                                  scale(BGB_Spawn)+ 
                                  scale(carbon_stock)+
                                  scale(river)+ 
                                  scale(NDVI)+ 
                                  scale(human_foot), data=umf_y_full_1)


# # fit list for detection
# fms1<-fitList("p(.) Ocu(.)"=fit_unm_sp2_0,
#               #"p(h) Ocu(.)"=fit_unm_sp2_0h, # NA problem!
#               #"p(c) Ocu(.)"=fit_unm_sp2_0c)#,
#               "p(.) Ocu(BGB_Spawn)"=fit_unm_sp2_1,
#               "p(.) Ocu(human_foot)"=fit_unm_sp2_2,
#               "p(.) Ocu(NDVI)"=fit_unm_sp2_3,
#               "p(.) Ocu(river)"=fit_unm_sp2_4,
#               "p() Ocu(carbon)"=fit_unm_sp2_5,
#               "p() Ocu(canopy)"=fit_unm_sp2_6,
#               )

# model selection detection unmarked
# modSel(fms1) #problem!.. null model ranks better

# fit list for occupancy
fms1<-fitList("p(.) Ocu(.)"=fit_unm_sp1_0,
              #"p(h) Ocu(.)"=fit_unm_sp2_0h, #,
              #"p(c) Ocu(.)"=fit_unm_sp2_0c,
              "p(.) Ocu(BGB_Spawn)"=fit_unm_sp1_1,
              "p(.) Ocu(human_foot)"=fit_unm_sp1_2,
              "p(.) Ocu(NDVI)"=fit_unm_sp1_3,
              "p(.) Ocu(river)"=fit_unm_sp1_4,
              #"p(.) Ocu(river^2)"=fit_unm_sp4_4s,
              "p() Ocu(carbon)"=fit_unm_sp1_5,
              "p() Ocu(canopy)"=fit_unm_sp1_6,
              "p() Ocu(canopy+BGB_Spawn)"=fit_unm_sp1_7,
              "p() Ocu(BGB_Spawn+human_foot)"=fit_unm_sp1_8,
              "p() Ocu(river+NDVI+human_foot)"=fit_unm_sp1_9,
              "p() Ocu(canopy+BGB_Spawn+carbon)"=fit_unm_sp1_10,
              "p() Ocu(ALL)"=fit_unm_sp1_11
              )

# model selection detection unmarked
modSel(fms1)

summary(fit_unm_sp1_1)

############# histogram evaluation
pb_f <- parboot(fit_unm_sp1_1, nsim=500, report=10) 
plot (pb_f)


pred_psi <-predict(fit_unm_sp1_1, type="state", newdata=newdat_range, appendData=TRUE) 


###  Plot occupancy en escala original
plot(Predicted~BGB_Spawn, pred_psi,type="l",col="blue", ylim=c(-0.1,1.1),
     xlab="BGB_Spawn",
     ylab="Probabilidad de ocupación")
lines(lower~BGB_Spawn, pred_psi,type="l",col=gray(0.5))
lines(upper~BGB_Spawn, pred_psi,type="l",col=gray(0.5))



```


### Spatial Prediction

```{r}
#| warning: false
#| message: false

#### predict

# srtm_crop_s <- mask(elev, NDVI)#stack(raster(elev))#, 
# scale(roughness)) # scale altitud
# names(srtm_crop_s) <- c("elevation")#, "roughness")
# crs(srtm_crop_s) <- "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0"
# nd1 <- NDVI #data.frame(srtm_crop_s, hour=0.5)
# names(nd1) <- "NDVI"
# newd <- data.frame(srtm_crop_s, hour=0.5)
# pred_psi_s_unm <-predict(fit_unm_sp1_5, type="state", newdata=nd1)
#pred_psi_s_stan <-predict(fit_stan_sp2_1, submodel="state", newdata=nd, na.rm=FALSE)
#pred_psi_ras <- raster(srtm_crop_s) # make raster = to elev 
#pred_psi_ras[] <- pred_psi_s_stan$Predicted # drape on top



newdata_terra = c(covs_raster[[1]])#,#,
                 #covs_raster[[6]])#,
                 #covs_raster[[6]])

occuPred <- terra::predict(newdata_terra, fit_unm_sp1_1, type="state", cores=3, na.rm=TRUE, progress = 'text')



# pred_psi_r <- pred_psi_s # * sd(full_covs$elevation) + mean(full_covs$elevation)
# crs(pred_psi_ras) <- "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0"
clr <- colorRampPalette(brewer.pal(9, "YlGn"))
# mapview (pred_psi_r[[1]], col.regions = clr,  legend = TRUE, alpha=0.7)
# plot(pred_psi_s[[1]], main="Occupancy")
levelplot(occuPred[[1]], par.settings = YlOrRdTheme(), margin=FALSE, main="Ocupancy")

pal = mapviewPalette("mapviewSpectralColors")
mapview(raster(occuPred[[1]]), map.types = c("Esri.WorldImagery"), alpha=0.5, col.regions = pal(10)) + mapview(puntos_sf)


```

## Información de la sesión en R.

```{r sesion, results='markup'}
print(sessionInfo(), locale = FALSE)
```


