---
title: "BFI predictors, Parita Bay, Panama"
subtitle: "A spatial modeling approach "
date: "`r Sys.Date()`"
author: 
  - name: Diego J. Lizcano
    orcid: https://orcid.org/0000-0002-9648-0576
  - name: Jorge Velásquez-Tibata
    orcid: https://orcid.org/0000-0002-7773-7348
license: CC BY-SA
toc: true
format: 
  html:
    theme: cosmo
    code-fold: true
    code-block-bg: "gray70"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "png",
  dev.args = list(type = "cairo-png"),
  fig.width = 7,
  fig.height = 5,
  fig.align = "center",
  eval = TRUE,
  echo = TRUE,
  warning = FALSE,
  error = FALSE,
  message = FALSE,
  cache=TRUE)
```

## Why spatial model?

A spatial linear model is a statistical linear model that incorporates spatial covariance among neighboring observations. Formally incorporating this spatial covariance generally yields models that more realistically represent spatial processes.

## Question

Which are the best spatial predictors for BFI?

## Set up analysis

Load libraries and set some options.

```{r set_up}
# set up


library(gt)
library(lubridate)
library(stringr)
library(readxl)
library(sf)

library(spmodel)
library(terra)
library(broom)
library(StepGWR)
library(mapview)
library(corrplot)

library(tidyverse)
library(tidymodels)


# setwd("C:/CodigoR/AudubonPanama/R/BFI")

options(scipen=99999)
options(max.print=99999)
options(stringsAsFactors=F)
```

## Load Data

```{r}


covs <- read_csv("C:/CodigoR/AudubonPanama/shp/sites_covs_parita_nona.csv") |> mutate(site=Name) 

BFI_site<- read.csv("C:/CodigoR/AudubonPanama/data/BFI_site.csv", header = TRUE) |> left_join(covs)
# convierte covs a puntos terra
puntos <- vect(BFI_site, geom=c("Longitude", "Latitude"), crs="EPSG:4326")
# convierte a sf
BFI_sf <- sf::st_as_sf(puntos)

```

# Correlation in rasters

```{r}
BGB_Spawn <- rast("C:/CodigoR/AudubonPanama/raster/BGB Spawn.tif")
AGB_Spawn <- rast("C:/CodigoR/AudubonPanama/raster/AGB Spawn.tif")
NDVI <- rast("C:/CodigoR/AudubonPanama/raster/S2_NDVI_median_v2.tif")
roads <- rast("C:/CodigoR/AudubonPanama/raster/roads_final_v2.tif")
carbon_stock <- rast("C:/CodigoR/AudubonPanama/raster/soil_organic_carbon_stock_0-30m.tif")
canopy <- rast("C:/CodigoR/AudubonPanama/raster/canopy_height_jetz2.tif")
human_foot <- rast("C:/CodigoR/AudubonPanama/raster/human_footprint.tif")
river <- rast("C:/CodigoR/AudubonPanama/raster/rivers_final_v2.tif")
coast <- rast("C:/CodigoR/AudubonPanama/raster/coast_final_v2.tif")

forest_integrity <- rast("C:/CodigoR/AudubonPanama/raster/forest_integrity_index.tif")
# make elevation equal
# srtm_projected <- projectRaster(srtm_crop, crs=projection(NDVI), method="ngb") 
# elev <- mask(resample(rast(srtm_projected), NDVI), NDVI)

# list of terras SpatRasters  
many_rasters <- list(BGB_Spawn, AGB_Spawn, human_foot, NDVI, river, canopy, roads, forest_integrity, coast)
# terra stack
covs_many_raster <- rast(many_rasters)
names(covs_many_raster) <- c("BGB_Spawn","AGB_Spawn", 
                        "human_foot", "NDVI", "river",  
                        "canopy", "roads", "forest_integrity",
                        "coast")



co <- layerCor(covs_many_raster, "pearson")
corrplot(co$correlation)


```

## Correlation in sites (points)

```{r}
# extract values from raster 
# covs_all1 <- terra::extract(covs_many_raster, puntos) 

#saveRDS(covs_all1, "C:/CodigoR/AudubonPanama/data/BFI/covs_all.RDS")
# save(covs_all, file = "C:/CodigoR/AudubonPanama/data/BFI/covs_all.Rda")
covs_all <- readRDS("C:/CodigoR/AudubonPanama/data/BFI/covs_all.RDS")
#load the rda file
# covs_all <- load(file = "C:/CodigoR/AudubonPanama/data/BFI/covs_all.Rda")
# write.csv(covs_all)

# covs_all$site <- puntos$site # ad site name

# change NA to 0 
# covs_all <- substr(covs_all, NA, 0) #r eplace in terra
covs_all[is.na(covs_all)] <- 0
M = cor(covs_all[,c(-1, -11)]) # removes ID and site
corrplot(M)

```

> layer removed: 

```{r}

many_rasters <- list(BGB_Spawn, AGB_Spawn, human_foot, NDVI, river, canopy, roads, forest_integrity, coast)
# terra stack
# # covs_many_raster <- terra::rast(many_rasters)
# names(covs_many_raster) <- c("BGB_Spawn", "AGB_Spawn",
#                         "human_foot", "NDVI", "river",  
#                         "canopy", "roads", "forest_integrity",
#                         "coast")
# 
# writeRaster(covs_many_raster, "C:/CodigoR/AudubonPanama/raster/covs_many_raster.tif", overwrite=TRUE)

covs_many_raster <- rast("C:/CodigoR/AudubonPanama/raster/covs_many_raster.tif")

# extract values from raster 
# covs_all <- terra::extract(covs_many_raster, puntos) 
# covs_all$site <- puntos$site # ad site name

# change NA to 0 
# covs_all <- substr(covs_all, NA, 0) #r eplace in terra
# covs_all[is.na(covs_all)] <- 0


```


## Which covariates are best?

We ran a stepwise procedure for Geographically Weighed Regression to eliminate any redundant variables. This is a hybrid spatial model that combines the variable selection capabilities of stepwise regression with the predictive power of Geographically Weighted Regression (GWR) model.

```{r}



data_site <- BFI_site |> dplyr::select(site, bfi_unscaled, Latitude, Longitude) |> left_join(covs_all)

data_site <- data_site[,c(-1,-5)] # remove Site and ID
data_site <- data_site |> relocate("Latitude", .after = last_col())
data_site <- data_site |> relocate("Longitude", .after = last_col()) |> dplyr::select(bfi_unscaled,BGB_Spawn,AGB_Spawn,human_foot,canopy,forest_integrity, coast, Latitude,Longitude)

StepGWR_exp<-StepGWR_exponential(data_site,
                                 0.5, 0.7,
                                 exponential_kernel)

```

The most important variables are: `r StepGWR_exp$Important_vars` 


## Multiple models using the most important variables 

```{r}

# convierte covs a puntos terra
puntos <- vect(data_site, geom=c("Longitude", "Latitude"), crs="EPSG:4326")
# convierte a sf
BFI_sf <- sf::st_as_sf(puntos)

spmods_Spawn1 <- splm(formula = bfi_unscaled ~ AGB_Spawn, data = BFI_sf,
              spcov_type = c("exponential"))

spmods_Spawn2 <- splm(formula = bfi_unscaled ~ BGB_Spawn, data = BFI_sf,
              spcov_type = c("exponential"))

spmods_Spawn3 <- splm(formula = bfi_unscaled ~ AGB_Spawn + scale(BGB_Spawn), data = BFI_sf,
              spcov_type = c("exponential"))

spmods_Spawn4 <- splm(formula = bfi_unscaled ~ BGB_Spawn + AGB_Spawn + forest_integrity, data = BFI_sf,
              spcov_type = c("exponential"))


glances(spmods_Spawn1, spmods_Spawn2, spmods_Spawn3, spmods_Spawn4)
summary(spmods_Spawn4)
loocv(spmods_Spawn4, cv_predict = TRUE, se.fit = TRUE)
augment(spmods_Spawn4)

# plot(spmods_Spawn4, which = c(7))


```




```{r}
m1 <- glm(bfi_unscaled ~  AGB_Spawn+BGB_Spawn +forest_integrity, data = BFI_sf, family = gaussian(link = "identity"))
library (sjPlot)
plot_model(m1)

library(ggeffects)
mydf <- predict_response(m1, terms = "AGB_Spawn")
plot(mydf)
mydf <- predict_response(m1, terms = "BGB_Spawn")
plot(mydf)
mydf <- predict_response(m1, terms = "forest_integrity")
plot(mydf)
```

## Predict BFI in map

```{r}
BGB_Spawn <- terra::rast("C:/CodigoR/AudubonPanama/raster/BGB Spawn.tif")
AGB_Spawn <- terra::rast("C:/CodigoR/AudubonPanama/raster/AGB Spawn.tif")
forest_integrity <- terra::rast("C:/CodigoR/AudubonPanama/raster/forest_integrity_index.tif")

# list of terras SpatRasters  
final_rasters <- list(BGB_Spawn, AGB_Spawn, forest_integrity)
# terra stack
final_covs_raster <- rast(final_rasters)
names(final_covs_raster) <- c("BGB_Spawn", "AGB_Spawn",  "forest_integrity")

# regular points
pointopred <- spatSample(final_covs_raster, 100000, "regular", cells=FALSE, xy=TRUE, na.rm=TRUE)

pointopred_data <- as.data.frame(as.points(as.points(final_covs_raster), multi=TRUE))
pointopred_geom <- as.data.frame(geom(as.points(as.points(final_covs_raster), multi=TRUE))) 

pointopred_data$x <- pointopred_geom$x
pointopred_data$y <- pointopred_geom$y


# convierte covs a puntos terra
# pointopredt <- vect(pointopred, geom=c("x", "y"), crs="EPSG:4326")
pointopredt <- vect(pointopred_data, geom=c("x", "y"), crs="EPSG:4326")

# convierte a sf
# pointopredsf <- sf::st_as_sf(pointopredt)
# makes prediction on top of vector... SLOW!!!!
pointopredt$prediction <- predict(spmods_Spawn4, newdata = pointopredt)
# augment(spmods_Spawn4, newdata = pointopredsf, interval = "prediction")
# mapapredicho <- rast(pointopredsf$prediction)

predgrid = terra::rast(pointopredt, nrow=2880, ncol=2274)
ncr = rasterize(pointopredt, predgrid, field="prediction")
# ncr[is.na(ncr)] = 0 # make na to 
mapview(ncr)#, zcol = c("prediction"))


```


## Sesion info

```{r}
print(sessionInfo(), locale = FALSE)
```



