---
title: "BFI predictors, Parita Bay, Panama"
subtitle: "A spatial modeling approach "
date: "`r Sys.Date()`"
author: 
  - name: Diego J. Lizcano
    orcid: https://orcid.org/0000-0002-9648-0576
  - name: Jorge Velásquez-Tibata
    orcid: https://orcid.org/0000-0002-7773-7348
license: CC BY-SA
toc: true
format: 
  html:
    theme: cosmo
    code-fold: true
    code-block-bg: "gray70"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "png",
  dev.args = list(type = "cairo-png"),
  fig.width = 7,
  fig.height = 5,
  fig.align = "center",
  eval = TRUE,
  echo = TRUE,
  warning = FALSE,
  error = FALSE,
  message = FALSE,
  cache=TRUE)
```

## Why spatial model?

A spatial linear model is a statistical linear model that incorporates spatial covariance among neighboring observations. Formally incorporating this spatial covariance generally yields models that more realistically represent spatial processes.

## Question

Which are the best spatial predictors for BFI?

## Set up analysis

Load libraries and set some options.

```{r set_up}
# set up

library(tidyverse)
library(gt)
library(lubridate)
library(stringr)
library(readxl)
library(sf)
library(tidymodels)
library(spmodel)
library(terra)
library(broom)
library(StepGWR)
library(mapview)

# setwd("C:/CodigoR/AudubonPanama/R/BFI")

options(scipen=99999)
options(max.print=99999)
options(stringsAsFactors=F)
```

## Load Data

```{r}


covs <- read_csv("C:/CodigoR/AudubonPanama/shp/sites_covs_parita_nona.csv") |> mutate(site=Name) 

BFI_site<- read.csv("C:/CodigoR/AudubonPanama/data/BFI_site.csv", header = TRUE) |> left_join(covs)
# convierte covs a puntos terra
puntos <- vect(BFI_site, geom=c("Longitude", "Latitude"), crs="EPSG:4326")
# convierte a sf
BFI_sf <- sf::st_as_sf(puntos)

```

## Which covariates are best?

We ran a stepwise procedure for Geographically Weighed Regression to eliminate any
redundant variables. This is a hybrid spatial model that combines the variable selection capabilities of stepwise regression with the predictive power of Geographically Weighted Regression (GWR) model.

```{r}
data_sp <- BFI_site |> select(bfi_unscaled, AGB_Spawn, BGB_Spawn, canopy_height_jetz2, coast_final_v2,  forest_integrity_index, Latitude, Longitude)

StepGWR_exp<-StepGWR_exponential(data_sp,0.25,0.8,exponential_kernel)

```

The most important variables are: `r StepGWR_exp$Important_vars` 


## Multiple models using the most important variables 

```{r}
spmods_Spawn1 <- splm(formula = bfi_unscaled ~ AGB_Spawn, data = BFI_sf,
              spcov_type = c("exponential"))

spmods_Spawn2 <- splm(formula = bfi_unscaled ~ BGB_Spawn, data = BFI_sf,
              spcov_type = c("exponential"))

spmods_Spawn3 <- splm(formula = bfi_unscaled ~ AGB_Spawn + BGB_Spawn, data = BFI_sf,
              spcov_type = c("exponential"))

spmods_Spawn4 <- splm(formula = bfi_unscaled ~ BGB_Spawn + AGB_Spawn + forest_integrity_index, data = BFI_sf,
              spcov_type = c("exponential"))


glances(spmods_Spawn1, spmods_Spawn2, spmods_Spawn3, spmods_Spawn4)
summary(spmods_Spawn4)
loocv(spmods_Spawn4, cv_predict = TRUE, se.fit = TRUE)
augment(spmods_Spawn4)



```

## Predict BFI in map


```{r}
BGB_Spawn <- rast("C:/CodigoR/AudubonPanama/raster/BGB Spawn.tif")
AGB_Spawn <- rast("C:/CodigoR/AudubonPanama/raster/AGB Spawn.tif")
forest_integrity_index <- rast("C:/CodigoR/AudubonPanama/raster/forest_integrity_index.tif")

# list of terras SpatRasters  
many_rasters <- list(BGB_Spawn, AGB_Spawn, forest_integrity_index)
# terra stack
covs_raster <- rast(many_rasters)
names(covs_raster) <- c("BGB_Spawn", "AGB_Spawn",  "forest_integrity_index")

# regular points
pointopred <- spatSample(covs_raster, 100000, "regular", cells=FALSE, xy=TRUE, na.rm=TRUE)

# convierte covs a puntos terra
pointopredt <- vect(pointopred, geom=c("x", "y"), crs="EPSG:4326")
# convierte a sf
pointopredsf <- sf::st_as_sf(pointopredt)

pointopredsf$prediction <- predict(spmods_Spawn4, newdata = pointopredsf)
# augment(spmods_Spawn4, newdata = pointopredsf, interval = "prediction")

mapview(pointopredsf, zcol = c("prediction"))


```


## Sesion info

```{r}
print(sessionInfo(), locale = FALSE)
```



