---
title: "Abundance all species as a function. Parita Bay, Panama"
subtitle: "site abundance prediction"
date: "`r Sys.Date()`"
author: 
  - name: Diego J. Lizcano
    orcid: https://orcid.org/0000-0002-9648-0576
  - name: Jorge Velásquez-Tibata
    orcid: https://orcid.org/0000-0002-7773-7348
license: CC BY-SA
toc: true
format: 
  html:
    theme: cosmo
    code-fold: true
    code-block-bg: "gray70"
---

Fit the occupancy model of Royle and Nichols (2003), which relates probability of detection of the species to the number of individuals available for detection at each site.

Royle, J. A. and Nichols, J. D. (2003) Estimating Abundance from Repeated Presence-Absence Data or Point Counts. Ecology, 84(3) pp. 777–790.

The number of animals available for detection at site _i_ is modelled as Poisson:
$$
N_1 \sim Poisson(\lambda_i)
$$

We assume that all individuals (of same species) at site _i_ during sample _j_ have identical detection probabilities, $ r_{i,j} $ and and that detections are independent. The species will be recorded if at least one individual is detected. Thus, the detection probability for the species is linked to the detection probability for an individual by:

$$
p_{i,j} = 1-(1-r_{i,j})^{N_i}
$$
The equation for the detection history is then:
$$
y_{i,j} \sim Bernoulli(p_{i,j})
$$

# Read data

## grouping by 3 days

```{r}
#| warning: false
#| message: false

# load function
source("C:/CodigoR/AudubonPanama/R/matrix_creator.R")
############### load Packages
library(readr) # readr 
library(readxl) # read excel
library(tidyverse) # 
library(unmarked) # occupancy by maximun likelihood
library(sf) # simple feature
library(mapview) # view map
library(ubms) # occupancy by Bayesian made easy
library(stocc) # spatial data for example
library(terra) 
library(raster)
library(RColorBrewer)
library(rasterVis)



############### load data
# old 
# parita_data_full <- read_delim("C:/CodigoR/AudubonPanama/data/all_records_audubon_panama_to_occu.csv", 
#                               delim = "\t", escape_double = FALSE, 
#                               col_types = cols(eventDate = col_date(format = "%Y-%m-%d")), 
#                               trim_ws = TRUE)

# new
parita_data_full <- read_excel("C:/CodigoR/AudubonPanama/data/all_records_audubon_panama_Darwin_Core_revision.xlsx", 
    col_types = c("text", "text", "text", 
        "date", "text", "text", "numeric", 
        "numeric", "text", "text", "numeric", 
        "text", "numeric", "numeric", "text", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "text", "text", 
        "numeric", "text", "numeric", "text", 
        "text", "text", "text", "text", "numeric", 
        "text", "numeric", "text", "numeric", 
        "text", "text"))

covs <- read_csv("C:/CodigoR/AudubonPanama/shp/sites_covs_parita_nona.csv")

## Nyctidromus albicollis
## Aramus guarauna
## Buteogallus anthracinus
## Dryocopus lineatus
## Piranga rubra

# filter by threshold and species
parita_data_allsp <-  parita_data_full %>% 
                      filter(confidence >= threshold)# %>%
                    # filter(scientificName == c("Nyctidromus albicollis", 
                    #                             "Aramus guarauna",
                    #                             "Buteogallus anthracinus",
                    #                             "Dryocopus lineatus",
                    #                             "Piranga rubra"
                    #                             ))  

# insert Max and min to get matrix limits
# start_date
parita_data2 <- parita_data_allsp %>% 
  group_by(locationID) %>% 
  mutate(start_date = min(eventDate),
         end_date= max(eventDate))

# apply function to get matrix per species
parita <- f.matrix.creator2(data = parita_data2, year = 2023)


# # centroide for terra
# # centoide <- centroids(puntos, TRUE)
# centroide <- c(mean(as.numeric(parita_data_allsp$decimalLongitude)), mean(as.numeric(parita_data_allsp$decimalLatitude)))
# clip_window <- extent(-80.67105, -80.05822, 7.713744, 8.489888) # 7.932311, -80.612233  8.360073, -80.180109
# bb <- c(-80.612233, -80.180109, 7.932311, 8.360073)
# srtm <- raster::getData('SRTM', lon=centroide[1], lat=centroide[2])

# crop the  raster using the vector extent
# srtm_crop <- raster::crop(srtm, clip_window)
 #rast(srtm_crop) # convert to terra


# altitud <- elevation_3s(-72.893262, 7.664081007, path="data")
# altitud <- as.numeric(Camptostoma_obsoletum_occu_dia$Altitude)

# convierte covs a puntos terra
puntos <- vect(covs, geom=c("Longitude", "Latitude"), crs="EPSG:4326")
# convierte a sf
puntos_sf <- sf::st_as_sf(puntos)

######### precipitacion
precip_Anton <- read_csv("C:/CodigoR/AudubonPanama/data/Precip_Anton.csv")
precip <-  data.frame(f.shrink.matrix.h.to9(precip_Anton))
precip[,1] <- 0 #fix start

# extract elev from raster
# cam_covs <- raster::extract(srtm_crop, puntos_sf)
# 
# BGB_Spawn <- rast("C:/CodigoR/AudubonPanama/raster/BGB Spawn.tif")
# AGB_Spawn <- rast("C:/CodigoR/AudubonPanama/raster/AGB Spawn.tif")
# NDVI <- rast("C:/CodigoR/AudubonPanama/raster/S2_NDVI_median_v2.tif")
# roads <- rast("C:/CodigoR/AudubonPanama/raster/roads_final_v2.tif")
# carbon_stock <- rast("C:/CodigoR/AudubonPanama/raster/soil_organic_carbon_stock_0-30m.tif")
# canopy <- rast("C:/CodigoR/AudubonPanama/raster/canopy_height_jetz2.tif")
# human_foot <- rast("C:/CodigoR/AudubonPanama/raster/human_footprint.tif")
# river <- rast("C:/CodigoR/AudubonPanama/raster/rivers_final_v2.tif")

# make elevation equal
# srtm_projected <- projectRaster(srtm_crop, crs=projection(NDVI), method="ngb") 
# elev <- mask(resample(rast(srtm_projected), NDVI), NDVI)



# # list of terras SpatRasters  
# many_rasters <- list(BGB_Spawn, human_foot, NDVI, river, carbon_stock, canopy)
# # terra stack
# covs_raster <- rast(many_rasters)
# names(covs_raster) <- c("BGB_Spawn", "human_foot",  "NDVI", "river", "carbon_stock",  "canopy")
# 
# # change NA to 0 
# covs_raster0 <- subst(covs_raster, NA, 0)
# # extract covariates
# covs_raster_val <- extract(covs_raster0, puntos_sf)
# 
# plot(covs_raster)

##########
# chk names of species
# dete es hora, occur es presencia en el dia
# names(parita$dete[]) 


```


# function to get abundance 

These methods return an object storing the posterior distributions of the latent variables at each site. Empirical Bayes methods used by `unmarked::ranef` underestimate the variance of the posterior distribution because they do not account for uncertainty in the hyperparameters (lambda). Note also that the posterior mode appears to exhibit some bias as an estimator or abundance. Consider using the posterior mean instead, even though it will not be an integer in general. 


### One species, single season

```{r}
#| warning: false
#| message: false

ranef_by_sp <- function(spx=1){ suppressWarnings({ #species name
# Make UMF object
umf_y_full_spx<- unmarkedFrameOccu(y=f.shrink.matrix.to9( 
                                   as.data.frame(parita$occur[[spx]])),
                                   siteCovs=NULL,
                                   obsCovs=list(prec=precip,
                            hour= data.frame(hour = f.shrink.matrix.h.to9(parita$dete[[spx]]))
                                                )
                                   ) 


# fit unmarked
fit_unm_spx_0 <- unmarked::occuRN(~1~1, data=umf_y_full_spx) # ok!
fit_unm_spx_p <- unmarked::occuRN(~ scale(prec) ~1, data=umf_y_full_spx) # ok!
fit_unm_spx_h <- unmarked::occuRN(~ scale(hour) ~1, data=umf_y_full_spx) # ok!


# fit list for occupancy
fms2<-fitList("p(.) lambda(.)"=fit_unm_spx_0,
              "p(prec) lambda(.)"=fit_unm_spx_p,
              "p(hour) lambda(.)"=fit_unm_spx_h)
              # #"p(h) Ocu(.)"=fit_unm_sp2_0h, #,
              # #"p(c) Ocu(.)"=fit_unm_sp2_0c,
              # "p(.) lambda(BGB_Spawn)"=fit_unm_sp2_1,
              # "p(.) lambda(human_foot)"=fit_unm_sp2_2,
              # "p(.) lambda(NDVI)"=fit_unm_sp2_3,
              # "p(.) lambda(river)"=fit_unm_sp2_4,
              # #"p(.) Ocu(river^2)"=fit_unm_sp2_4s,
              # "p() lambda(carbon)"=fit_unm_sp2_5,
              # "p() lambda(canopy)"=fit_unm_sp2_6,
              # "p() lambda(canopy+BGB_Spawn)"=fit_unm_sp2_7,
              # "p() lambda(BGB_Spawn+human_foot)"=fit_unm_sp2_8,
              # "p() lambda(ALL)"=fit_unm_sp2_9
              # )

# model selection detection unmarked
 models <- modSel(fms2)
 models
 

# Empirical Bayes estimates of abundance at each site by species
re_0 <- ranef(fit_unm_spx_0)
re_h <- ranef(fit_unm_spx_h)
# plot(re)
# post.df <- as(re_0, "data.frame")

#####################################################
# to save the ranef result (S3 Object) as dataframe #
#####################################################

abund_per_site <- data.frame(sp=names(parita$dete[spx]), 
                             site=names( parita$occu[[1]][,1]),
                             mode=(bup(re_0, stat="mode") ),
                             mean=(bup(re_0, stat="mean") ), 
                             confint(re_0, level=0.9))

}) # no warning part
  
  
# return(list(paste("###", sp=names(parita$dete[spx]) ), # sp name
#             models, # model AICs
#             Null=re_0, # mull model estimates
#             hour=re_h)) # hour in detection estimates

  # return to export table
  return(abund_per_site)
  
} # end of function


```



## use the function

Runs a loop on a list containing the number of the species.  

```{r }

# list
abun_site <- list()

for(i in 1:length(names(parita$dete[]))){
  # print( ranef_by_sp(i) )
  print(paste(i, names(parita$dete[i])))
   abun_site[[i]] <-  ranef_by_sp(i) 
  
}

library (plyr)
# Convert the list to a dataframe
df <- ldply(abun_site, data.frame)
#save to table
write.csv(df, "C:/CodigoR/AudubonPanama/data/abundance_result.csv")


```



## Information of R session.

```{r sesion, results='markup'}
print(sessionInfo(), locale = FALSE)
```


