---
title: "Occupancy five species. Parita Bay, Panama"
subtitle: "spatial prediction"
date: "`r Sys.Date()`"
author: 
  - name: Diego J. Lizcano
    orcid: https://orcid.org/0000-0002-9648-0576
  - name: Jorge Velásquez-Tibata
    orcid: https://orcid.org/0000-0002-7773-7348
license: CC BY-SA
toc: true
format: 
  html:
    theme: cosmo
    code-fold: true
    code-block-bg: "gray70"
---


# Read spatial layers

```{r}
#| warning: false
#| message: false

# load function
source("C:/CodigoR/AudubonPanama/R/matrix_creator.R")
############### load Packages
library(readr) # readr 
library(readxl) # read excel
library(tidyverse) # 
library(unmarked) # occupancy by maximun likelihood
library(sf) # simple feature
library(mapview) # view map
library(ubms) # occupancy by Bayesian made easy
library(stocc) # spatial data for example
library(terra) 
library(raster)
library(RColorBrewer)
library(rasterVis)


############### load data
# old 
# parita_data_full <- read_delim("C:/CodigoR/AudubonPanama/data/all_records_audubon_panama_to_occu.csv", 
#                               delim = "\t", escape_double = FALSE, 
#                               col_types = cols(eventDate = col_date(format = "%Y-%m-%d")), 
#                               trim_ws = TRUE)

# new
parita_data_full <- read_excel("C:/CodigoR/AudubonPanama/data/all_records_audubon_panama_Darwin_Core_revision.xlsx", 
    col_types = c("text", "text", "text", 
        "date", "text", "text", "numeric", 
        "numeric", "text", "text", "numeric", 
        "text", "numeric", "numeric", "text", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "text", "text", 
        "numeric", "text", "numeric", "text", 
        "text", "text", "text", "text", "numeric", 
        "text", "numeric", "text", "numeric", 
        "text", "text"))

covs <- read_csv("C:/CodigoR/AudubonPanama/shp/sites_covs_parita_nona.csv")

## Nyctidromus albicollis
## Aramus guarauna
## Buteogallus anthracinus
## Dryocopus lineatus
## Piranga rubra

# filter by threshold and species
parita_data_5sp <-  parita_data_full %>% 
                      filter(confidence >= threshold) %>%
                      filter(scientificName == c("Nyctidromus albicollis", 
                                                 "Aramus guarauna",
                                                 "Buteogallus anthracinus",
                                                 "Dryocopus lineatus",
                                                 "Piranga rubra"
                                                 ))  

# insert Max and min to get matrix limits
# start_date
parita_data2 <- parita_data_5sp %>% 
  group_by(locationID) %>% 
  mutate(start_date = min(eventDate),
         end_date= max(eventDate))

# apply function to get matrix per species
parita <- f.matrix.creator2(data = parita_data2, year = 2023)


# centroide for terra
# centoide <- centroids(puntos, TRUE)
centroide <- c(mean(as.numeric(parita_data_5sp$decimalLongitude)), mean(as.numeric(parita_data_5sp$decimalLatitude)))
clip_window <- extent(-80.67105, -80.05822, 7.713744, 8.489888) # 7.932311, -80.612233  8.360073, -80.180109
bb <- c(-80.612233, -80.180109, 7.932311, 8.360073)
srtm <- raster::getData('SRTM', lon=centroide[1], lat=centroide[2])

# crop the  raster using the vector extent
srtm_crop <- raster::crop(srtm, clip_window)
 #rast(srtm_crop) # convert to terra


# altitud <- elevation_3s(-72.893262, 7.664081007, path="data")
# altitud <- as.numeric(Camptostoma_obsoletum_occu_dia$Altitude)

# convierte covs a puntos terra
puntos <- vect(covs, geom=c("Longitude", "Latitude"), crs="EPSG:4326")
# convierte a sf
puntos_sf <- sf::st_as_sf(puntos)
# extract elev from raster
cam_covs <- raster::extract(srtm_crop, puntos_sf)


AGB_Spawn <- rast("C:/CodigoR/AudubonPanama/raster/AGB Spawn.tif")
NDVI <- rast("C:/CodigoR/AudubonPanama/raster/S2_NDVI_median_v2.tif")
roads <- rast("C:/CodigoR/AudubonPanama/raster/roads_final_v2.tif")
carbon_stock <- rast("C:/CodigoR/AudubonPanama/raster/soil_organic_carbon_stock_0-30m.tif")
canopy <- rast("C:/CodigoR/AudubonPanama/raster/canopy_height_jetz2.tif")
# make elevation equal
srtm_projected <- projectRaster(srtm_crop, crs=projection(NDVI), method="ngb") 
elev <- mask(resample(rast(srtm_projected), NDVI), NDVI)



# list of terras SpatRasters  
many_rasters <- list(AGB_Spawn, NDVI, roads, carbon_stock, elev, canopy)
# terra stack
covs_raster <- rast(many_rasters)
names(covs_raster) <- c("AGB_Spawn", "NDVI", "roads", "carbon_stock", "elevation", "canopy")

# change NA to 0 
covs_raster0 <- subst(covs_raster, NA, 0)
# extract covariates
covs_raster_val <- extract(covs_raster0, puntos_sf)

plot(covs_raster)

##########
# chk names of species
# dete es hora, occur es presencia en el dia
# names(parita$dete[]) 


```


# *Buteogallus anthracinus*

### One species, single season

```{r}
#| warning: false
#| message: false

# Make UMF object
umf_y_full_2<- unmarkedFrameOccu(y= as.data.frame(parita$occur[[2]])) #"Aramus guarauna"
siteCovs(umf_y_full_2) <- data.frame(elevation =  covs_raster_val$elevation,
                                   AGB_Spawn = covs_raster_val$AGB_Spawn,
                                   roads = covs_raster_val$roads,
                                   carbon = covs_raster_val$carbon_stock,
                                   NDVI = covs_raster_val$NDVI,
                                   canopy = covs_raster_val$canopy
) # data.frame(Elev=full_covs$Elev) # Full


plot(umf_y_full_2, main="Buteogallus anthracinus")  #""Buteogallus anthracinus 

# obscanop<-matrix(rep(covs_raster_val$canopy_height_jetz2,27),ncol = 27) # make elevetion per observation

# detection
obsCovs(umf_y_full_2) <- list(hour = matrix(parita$dete[[2]]),
                              canopy = covs_raster_val$canopy ) # hora


# fit unmarked
fit_unm_sp2_0 <- unmarked::occu(~1~1, data=umf_y_full_2) # ok!
fit_unm_sp2_0h <- unmarked::occu(~scale(hour)~1, data=umf_y_full_2) # It work!
fit_unm_sp2_0c <- unmarked::occu(~scale(canopy)~1, data=umf_y_full_2) # It work!
fit_unm_sp2_1 <- unmarked::occu(~1~scale(elevation), data=umf_y_full_2)
fit_unm_sp2_2 <- unmarked::occu(~1~scale(AGB_Spawn), data=umf_y_full_2)
fit_unm_sp2_3 <- unmarked::occu(~1~scale(roads), data=umf_y_full_2) # ok!
fit_unm_sp2_4 <- unmarked::occu(~1~scale(carbon), data=umf_y_full_2)
fit_unm_sp2_5 <- unmarked::occu(~1~scale(NDVI), data=umf_y_full_2)
fit_unm_sp2_6 <- unmarked::occu(~1~scale(canopy), data=umf_y_full_2)


# fit list for detection
fms1<-fitList("p(.) Ocu(.)"=fit_unm_sp2_0,
              #"p(h) Ocu(.)"=fit_unm_sp2_0h, # NA problem!
              "p(c) Ocu(.)"=fit_unm_sp2_0c)#,
#              "p(.) Ocu(elev)"=fit_unm_sp2_1,
#              "p(.) Ocu(slope)"=fit_unm_sp2_2,
#              "p(.) Ocu(aspect)"=fit_unm_sp2_3,
#              "p(.) Ocu(roughness)"=fit_unm_sp2_4,
#              "p(elev^2) Ocu(elev^2)"=fit_unm_sp2_5 )

# model selection detection unmarked
modSel(fms1) #problem!.. null model ranks better

# fit list for detection
fms2<-fitList("p(.) Ocu(.)"=fit_unm_sp2_0,
              #"p(h) Ocu(.)"=fit_unm_sp2_0h, #,
              #"p(c) Ocu(.)"=fit_unm_sp2_0c,
              "p(.) Ocu(elev)"=fit_unm_sp2_1,
              "p(.) Ocu(AGB_Spawn)"=fit_unm_sp2_2,
              "p(.) Ocu(roads)"=fit_unm_sp2_3,
              "p(.) Ocu(carbon)"=fit_unm_sp2_4,
              "p(.) Ocu(NDVI)"=fit_unm_sp2_5,
              "p(.) Ocu(canopy)"=fit_unm_sp2_6)

# model selection detection unmarked
modSel(fms2)

######## ubms, bayesian occupancy
# Double formula: first part is for detection, second for occupancy
#  form <- ~detCov1 + detCov2 ~habCov1 + habCov2 + RSR(x, y, threshold=1)

# fit stan
fit_stan_sp2_0 <- stan_occu(~1~1, data=umf_y_full_2, chains=3, iter=500, cores=3)
fit_stan_sp2_0h <- stan_occu(~scale(hour)~1, data=umf_y_full_2, chains=3, iter=500, cores=3)
fit_stan_sp2_0c <- stan_occu(~scale(canopy)~1, data=umf_y_full_2, chains=3, iter=500, cores=3)
fit_stan_sp2_1 <- stan_occu(~scale(hour)~scale(elevation), data=umf_y_full_2, chains=3, iter=500, cores=3)
fit_stan_sp2_2 <- stan_occu(~scale(hour)~scale(AGB_Spawn), data=umf_y_full_2, chains=3, iter=500, cores=3)
fit_stan_sp2_3 <- stan_occu(~scale(hour)~scale(roads), data=umf_y_full_2, chains=3, iter=500, cores=3)
fit_stan_sp2_4 <- stan_occu(~scale(hour)~scale(carbon), data=umf_y_full_2, chains=3, iter=500, cores=3)
fit_stan_sp2_5 <- stan_occu(~scale(hour)~scale(NDVI), data=umf_y_full_2, chains=3, iter=500, cores=3)
fit_stan_sp2_6 <- stan_occu(~scale(hour)~scale(canopy), data=umf_y_full_2, chains=3, iter=500, cores=3)


mods2 <- fitList(fit_stan_sp2_0,
                fit_stan_sp2_0h,
                fit_stan_sp2_0c,
                fit_stan_sp2_1,
                fit_stan_sp2_2,
                fit_stan_sp2_3,
                fit_stan_sp2_4,
                fit_stan_sp2_5,
                fit_stan_sp2_6)

# model selection
round(modSel(mods2), 3)

# plot top model
print("Hour")
plot_effects(fit_stan_sp2_5, "det") # Detection
print("NDVI")
plot_effects(fit_stan_sp2_5, "state") # Occupancy

```

### Spatial Prediction

```{r}
#| warning: false
#| message: false

#### predict


# srtm_crop_s <- mask(elev, NDVI)#stack(raster(elev))#, 
#scale(roughness)) # scale altitud
# names(srtm_crop_s) <- c("elevation")#, "roughness")
# crs(srtm_crop_s) <- "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0"
nd2 <- NDVI # mask(elev, NDVI) #data.frame(srtm_crop_s, hour=0.5)
names(nd2) <- "NDVI"
# newd <- data.frame(srtm_crop_s, hour=0.5)
pred_psi_s_unm <-predict(fit_unm_sp2_5, type="state", newdata=nd2)
#pred_psi_s_stan <-predict(fit_stan_sp2_1, submodel="state", newdata=nd, na.rm=FALSE)
#pred_psi_ras <- raster(srtm_crop_s) # make raster = to elev 
#pred_psi_ras[] <- pred_psi_s_stan$Predicted # drape on top


# pred_psi_r <- pred_psi_s # * sd(full_covs$elevation) + mean(full_covs$elevation)
# crs(pred_psi_ras) <- "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0"
clr <- colorRampPalette(brewer.pal(9, "YlGn"))
# mapview (pred_psi_r[[1]], col.regions = clr,  legend = TRUE, alpha=0.7)
# plot(pred_psi_s[[1]], main="Occupancy")
levelplot(pred_psi_s_unm[[1]], par.settings = YlOrRdTheme(), margin=FALSE, main="Ocupancy")

pal = mapviewPalette("mapviewSpectralColors")
mapview(raster(pred_psi_s_unm[[1]]), map.types = c("Esri.WorldImagery"), alpha=0.5, col.regions = pal(10)) + mapview(puntos_sf)


```


# *Dryocopus lineatus*

### One species, single season


```{r}
#| warning: false
#| message: false

# Make UMF object
umf_y_full_3<- unmarkedFrameOccu(y= as.data.frame(parita$occur[[3]])) #Dryocopus lineatus
siteCovs(umf_y_full_3) <- data.frame(elevation =  covs_raster_val$elevation,
                                   AGB_Spawn = covs_raster_val$AGB_Spawn,
                                   roads = covs_raster_val$roads,
                                   carbon = covs_raster_val$carbon_stock,
                                   NDVI = covs_raster_val$NDVI,
                                   canopy = covs_raster_val$canopy
) # data.frame(Elev=full_covs$Elev) # Full


plot(umf_y_full_3, main="Dryocopus lineatus")  #""Buteogallus anthracinus no converge

# obscanop<-matrix(rep(covs_raster_val$canopy_height_jetz2,27),ncol = 27) # make elevetion per observation

# detection
obsCovs(umf_y_full_3) <- list(hour = matrix(parita$dete[[3]]),
                              canopy = covs_raster_val$canopy ) # hora


# fit unmarked
fit_unm_sp3_0 <- unmarked::occu(~1~1, data=umf_y_full_3) # ok!
fit_unm_sp3_0h <- unmarked::occu(~scale(hour)~1, data=umf_y_full_3) # It work!
fit_unm_sp3_0c <- unmarked::occu(~scale(canopy)~1, data=umf_y_full_3) # It work!
fit_unm_sp3_1 <- unmarked::occu(~1~scale(elevation), data=umf_y_full_3)
fit_unm_sp3_2 <- unmarked::occu(~1~scale(AGB_Spawn), data=umf_y_full_3)
fit_unm_sp3_3 <- unmarked::occu(~1~scale(roads), data=umf_y_full_3) # ok!
fit_unm_sp3_4 <- unmarked::occu(~1~scale(carbon), data=umf_y_full_3)
fit_unm_sp3_5 <- unmarked::occu(~1~scale(NDVI), data=umf_y_full_3)
fit_unm_sp3_6 <- unmarked::occu(~1~scale(canopy), data=umf_y_full_3)


# fit list for detection
fms1<-fitList("p(.) Ocu(.)"=fit_unm_sp3_0,
              # "p(h) Ocu(.)"=fit_unm_sp3_0h, # NA problem!
              "p(c) Ocu(.)"=fit_unm_sp3_0c)#,
#              "p(.) Ocu(elev)"=fit_unm_sp2_1,
#              "p(.) Ocu(slope)"=fit_unm_sp2_2,
#              "p(.) Ocu(aspect)"=fit_unm_sp2_3,
#              "p(.) Ocu(roughness)"=fit_unm_sp2_4,
#              "p(elev^2) Ocu(elev^2)"=fit_unm_sp2_5 )

# model selection detection unmarked
modSel(fms1) #problem!.. null model ranks better

# fit list for detection
fms2<-fitList("p(.) Ocu(.)"=fit_unm_sp3_0,
              #"p(h) Ocu(.)"=fit_unm_sp2_0h, #,
              #"p(c) Ocu(.)"=fit_unm_sp3_0c,
              "p(.) Ocu(elev)"=fit_unm_sp3_1,
              "p(.) Ocu(AGB_Spawn)"=fit_unm_sp3_2,
              "p(.) Ocu(roads)"=fit_unm_sp3_3,
              "p(.) Ocu(carbon)"=fit_unm_sp3_4,
              "p(.) Ocu(NDVI)"=fit_unm_sp3_5,
              "p(.) Ocu(canopy)"=fit_unm_sp3_6)

# model selection detection unmarked
modSel(fms2)

# fit stan
fit_stan_sp3_0 <- stan_occu(~1~1, data=umf_y_full_3, chains=3, iter=500, cores=3)
fit_stan_sp3_0h <- stan_occu(~scale(hour)~1, data=umf_y_full_3, chains=3, iter=500, cores=3)
fit_stan_sp3_1 <- stan_occu(~scale(hour)~scale(elevation), data=umf_y_full_3, chains=3, iter=500, cores=3)
fit_stan_sp3_2 <- stan_occu(~scale(hour)~scale(AGB_Spawn), data=umf_y_full_3, chains=3, iter=500, cores=3)
fit_stan_sp3_3 <- stan_occu(~scale(hour)~scale(roads), data=umf_y_full_3, chains=3, iter=500, cores=3)
fit_stan_sp3_4 <- stan_occu(~scale(hour)~scale(carbon), data=umf_y_full_3, chains=3, iter=500, cores=3)
fit_stan_sp3_5 <- stan_occu(~scale(hour)~scale(NDVI), data=umf_y_full_3, chains=3, iter=500, cores=3)
fit_stan_sp3_6 <- stan_occu(~scale(hour)~scale(canopy), data=umf_y_full_3, chains=3, iter=500, cores=3)


mods3 <- fitList(fit_stan_sp3_0,
                fit_stan_sp3_0h,
                fit_stan_sp3_1,
                fit_stan_sp3_2,
                fit_stan_sp3_3,
                fit_stan_sp3_4,
                fit_stan_sp3_5,
                fit_stan_sp3_6)

# model selection
round(modSel(mods3), 3) #AGB_Spawn

# plot top model
print("Hour")
plot_effects(fit_stan_sp3_5, "det") # Detectiom
print("NDVI")
plot_effects(fit_stan_sp3_5, "state") # Ocupancy

```


### Spatial Prediction

```{r}
#| warning: false
#| message: false

#### predict

# srtm_crop_s <- mask(elev, NDVI)#stack(raster(elev))#, 
# scale(roughness)) # scale altitud
# names(srtm_crop_s) <- c("elevation")#, "roughness")
# crs(srtm_crop_s) <- "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0"
nd3 <- NDVI #mask(AGB_Spawn, NDVI) #data.frame(srtm_crop_s, hour=0.5)
names(nd3) <- "NDVI"
# newd <- data.frame(srtm_crop_s, hour=0.5)
pred_psi_s_unm <-predict(fit_unm_sp3_5, type="state", newdata=nd3)
#pred_psi_s_stan <-predict(fit_stan_sp2_1, submodel="state", newdata=nd, na.rm=FALSE)
#pred_psi_ras <- raster(srtm_crop_s) # make raster = to elev 
#pred_psi_ras[] <- pred_psi_s_stan$Predicted # drape on top


# pred_psi_r <- pred_psi_s # * sd(full_covs$elevation) + mean(full_covs$elevation)
# crs(pred_psi_ras) <- "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0"
clr <- colorRampPalette(brewer.pal(9, "YlGn"))
# mapview (pred_psi_r[[1]], col.regions = clr,  legend = TRUE, alpha=0.7)
# plot(pred_psi_s[[1]], main="Occupancy")
levelplot(pred_psi_s_unm[[1]], par.settings = YlOrRdTheme(), margin=FALSE, main="Ocupancy")

pal = mapviewPalette("mapviewSpectralColors")
mapview(raster(pred_psi_s_unm[[1]]), map.types = c("Esri.WorldImagery"), alpha=0.5, col.regions = pal(10)) + mapview(puntos_sf)


```



# *Nyctidromus albicollis*

### One species, single season


```{r}
#| warning: false
#| message: false

# Make UMF object
umf_y_full_4<- unmarkedFrameOccu(y= as.data.frame(parita$occur[[4]])) #Dryocopus lineatus
siteCovs(umf_y_full_4) <- data.frame(elevation =  covs_raster_val$elevation,
                                   AGB_Spawn = covs_raster_val$AGB_Spawn,
                                   roads = covs_raster_val$roads,
                                   carbon = covs_raster_val$carbon_stock,
                                   NDVI = covs_raster_val$NDVI,
                                   canopy = covs_raster_val$canopy
) # data.frame(Elev=full_covs$Elev) # Full


plot(umf_y_full_4, main="Nyctidromus albicollis")  #""Buteogallus anthracinus no converge

# obscanop<-matrix(rep(covs_raster_val$canopy_height_jetz2,27),ncol = 27) # make elevetion per observation

# detection
obsCovs(umf_y_full_4) <- list(hour = matrix(parita$dete[[4]]),
                              canopy = covs_raster_val$canopy ) # hora


# fit unmarked
fit_unm_sp4_0 <- unmarked::occu(~1~1, data=umf_y_full_4) # ok!
fit_unm_sp4_0h <- unmarked::occu(~scale(hour)~1, data=umf_y_full_4) # It work!
fit_unm_sp4_0c <- unmarked::occu(~scale(canopy)~1, data=umf_y_full_4) # It work!
fit_unm_sp4_1 <- unmarked::occu(~scale(canopy)~scale(elevation), data=umf_y_full_4)
fit_unm_sp4_2 <- unmarked::occu(~scale(canopy)~scale(AGB_Spawn), data=umf_y_full_4)
fit_unm_sp4_3 <- unmarked::occu(~scale(canopy)~scale(roads), data=umf_y_full_4) # ok!
fit_unm_sp4_4 <- unmarked::occu(~scale(canopy)~scale(carbon), data=umf_y_full_4)
fit_unm_sp4_5 <- unmarked::occu(~scale(canopy)~scale(NDVI), data=umf_y_full_4)
fit_unm_sp4_6 <- unmarked::occu(~scale(canopy)~scale(canopy), data=umf_y_full_4)
fit_unm_sp4_2b <- unmarked::occu(~1~scale(AGB_Spawn), data=umf_y_full_4)



# fit list for detection
fms1<-fitList("p(.) Ocu(.)"=fit_unm_sp4_0,
              # "p(h) Ocu(.)"=fit_unm_sp3_0h, # NA problem!
              "p(c) Ocu(.)"=fit_unm_sp4_0c)#,
#              "p(.) Ocu(elev)"=fit_unm_sp2_1,
#              "p(.) Ocu(slope)"=fit_unm_sp2_2,
#              "p(.) Ocu(aspect)"=fit_unm_sp2_3,
#              "p(.) Ocu(roughness)"=fit_unm_sp2_4,
#              "p(elev^2) Ocu(elev^2)"=fit_unm_sp2_5 )

# model selection detection unmarked
modSel(fms1) #good!.. canopy  model ranks better

# fit list for detection
fms2<-fitList("p(.) Ocu(.)"=fit_unm_sp4_0,
              #"p(h) Ocu(.)"=fit_unm_sp2_0h, #,
              "p(c) Ocu(.)"=fit_unm_sp4_0c,
              "p(c) Ocu(elev)"=fit_unm_sp4_1,
              "p(c) Ocu(AGB_Spawn)"=fit_unm_sp4_2,
              "p(c) Ocu(roads)"=fit_unm_sp4_3,
              "p(c) Ocu(carbon)"=fit_unm_sp4_4,
              "p(c) Ocu(NDVI)"=fit_unm_sp4_5,
              "p(c) Ocu(canopy)"=fit_unm_sp4_6)

# model selection detection unmarked
modSel(fms2)

# fit stan
fit_stan_sp4_0 <- stan_occu(~1~1, data=umf_y_full_4, chains=3, iter=500, cores=3)
fit_stan_sp4_0h <- stan_occu(~scale(hour)~1, data=umf_y_full_4, chains=3, iter=500, cores=3)
fit_stan_sp4_0c <- stan_occu(~scale(canopy)~1, data=umf_y_full_4, chains=3, iter=500, cores=3)
fit_stan_sp4_1 <- stan_occu(~1~scale(elevation), data=umf_y_full_4, chains=3, iter=500, cores=3)
fit_stan_sp4_2 <- stan_occu(~scale(hour)~scale(AGB_Spawn), data=umf_y_full_4, chains=3, iter=500, cores=3)
fit_stan_sp4_3 <- stan_occu(~scale(hour)~scale(roads), data=umf_y_full_4, chains=3, iter=500, cores=3)
fit_stan_sp4_4 <- stan_occu(~scale(hour)~scale(carbon), data=umf_y_full_4, chains=3, iter=500, cores=3)
fit_stan_sp4_5 <- stan_occu(~scale(hour)~scale(NDVI), data=umf_y_full_4, chains=3, iter=500, cores=3)


mods4 <- fitList(fit_stan_sp4_0,
                fit_stan_sp4_0h,
                fit_stan_sp4_0c,
                fit_stan_sp4_1,
                fit_stan_sp4_2,
                fit_stan_sp4_3,
                fit_stan_sp4_4,
                fit_stan_sp4_5)

# model selection stan
round(modSel(mods4), 3)

# plot top model
print("Hour")
plot_effects(fit_stan_sp4_2, "det") # Detection
print("AGB_Spawn")
plot_effects(fit_stan_sp4_2, "state") # Occupancy

```


### Spatial Prediction

```{r}
#| warning: false
#| message: false

#### predict

# srtm_crop_s <- mask(elev, NDVI)#stack(raster(elev))#, 
# scale(roughness)) # scale altitud
# names(srtm_crop_s) <- c("elevation")#, "roughness")
# crs(srtm_crop_s) <- "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0"
nd4 <- mask(AGB_Spawn, NDVI) #data.frame(srtm_crop_s, hour=0.5)
names(nd4) <- "AGB_Spawn"
# newd <- data.frame(srtm_crop_s, hour=0.5)
pred_psi_s_unm <-predict(fit_unm_sp4_2b, type="state", newdata=nd4)
#pred_psi_s_stan <-predict(fit_stan_sp2_1, submodel="state", newdata=nd, na.rm=FALSE)
#pred_psi_ras <- raster(srtm_crop_s) # make raster = to elev 
#pred_psi_ras[] <- pred_psi_s_stan$Predicted # drape on top


# pred_psi_r <- pred_psi_s # * sd(full_covs$elevation) + mean(full_covs$elevation)
# crs(pred_psi_ras) <- "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0"
clr <- colorRampPalette(brewer.pal(9, "YlGn"))
# mapview (pred_psi_r[[1]], col.regions = clr,  legend = TRUE, alpha=0.7)
# plot(pred_psi_s[[1]], main="Occupancy")
levelplot(pred_psi_s_unm[[1]], par.settings = YlOrRdTheme(), margin=FALSE, main="Ocupancy")

pal = mapviewPalette("mapviewSpectralColors")
mapview(raster(pred_psi_s_unm[[1]]), map.types = c("Esri.WorldImagery"), alpha=0.5, col.regions = pal(10)) + mapview(puntos_sf)


```




# *Piranga rubra*

### One species, single season


```{r}
#| warning: false
#| message: false

# Make UMF object
umf_y_full_5<- unmarkedFrameOccu(y= as.data.frame(parita$occur[[5]])) #Dryocopus lineatus
siteCovs(umf_y_full_5) <- data.frame(elevation =  covs_raster_val$elevation,
                                   AGB_Spawn = covs_raster_val$AGB_Spawn,
                                   roads = covs_raster_val$roads,
                                   carbon = covs_raster_val$carbon_stock,
                                   NDVI = covs_raster_val$NDVI,
                                   canopy = covs_raster_val$canopy
) # data.frame(Elev=full_covs$Elev) # Full


plot(umf_y_full_5, main="Piranga rubra")  #""Buteogallus anthracinus no converge

# detection
obsCovs(umf_y_full_5) <- list(hour = matrix(parita$dete[[5]]),
                              canopy = covs_raster_val$canopy ) # hora

# Double formula: first part is for detection, second for occupancy
#  form <- ~detCov1 + detCov2 ~habCov1 + habCov2 + RSR(x, y, threshold=1)

# fit unmarked
fit_unm_sp5_0 <- unmarked::occu(~1~1, data=umf_y_full_5) # ok!
# fit_unm_sp4_0h <- unmarked::occu(~hour~1, data=umf_y_full_4) # Does not work!
fit_unm_sp5_0c <- unmarked::occu(~scale(canopy)~1, data=umf_y_full_5) # Does not work!
fit_unm_sp5_1 <- unmarked::occu(~scale(canopy)~scale(elevation), data=umf_y_full_5)
fit_unm_sp5_2 <- unmarked::occu(~scale(canopy)~scale(AGB_Spawn), data=umf_y_full_5)
fit_unm_sp5_3 <- unmarked::occu(~scale(canopy)~scale(roads), data=umf_y_full_5) # ok!
fit_unm_sp5_4 <- unmarked::occu(~scale(canopy)~scale(carbon), data=umf_y_full_5)
fit_unm_sp5_5 <- unmarked::occu(~scale(canopy)~scale(NDVI), data=umf_y_full_5)
fit_unm_sp5_2b <- unmarked::occu(~1~scale(AGB_Spawn), data=umf_y_full_5)


# fit list for detection
fms1<-fitList("p(.) Ocu(.)"=fit_unm_sp5_0,
              # "p(h) Ocu(.)"=fit_unm_sp3_0h, # NA problem!
              "p(c) Ocu(.)"=fit_unm_sp5_0c)#,
#              "p(.) Ocu(elev)"=fit_unm_sp2_1,
#              "p(.) Ocu(slope)"=fit_unm_sp2_2,
#              "p(.) Ocu(aspect)"=fit_unm_sp2_3,
#              "p(.) Ocu(roughness)"=fit_unm_sp2_4,
#              "p(elev^2) Ocu(elev^2)"=fit_unm_sp2_5 )

# model selection detection unmarked
modSel(fms1) #good!.. canopy  model ranks better

# fit list for detection
fms2<-fitList("p(.) Ocu(.)"=fit_unm_sp4_0,
              #"p(h) Ocu(.)"=fit_unm_sp2_0h, #,
              "p(c) Ocu(.)"=fit_unm_sp4_0c,
              "p(c) Ocu(elev)"=fit_unm_sp4_1,
              "p(c) Ocu(AGB_Spawn)"=fit_unm_sp4_2,
              "p(c) Ocu(roads)"=fit_unm_sp4_3,
              "p(c) Ocu(carbon)"=fit_unm_sp4_4,
              "p(c) Ocu(NDVI)"=fit_unm_sp4_5,
              "p(c) Ocu(canopy)"=fit_unm_sp4_6)

# model selection detection unmarked
modSel(fms2)


# fit stan
fit_stan_sp5_0 <- stan_occu(~1~1, data=umf_y_full_5, chains=3, iter=500, cores=3)
fit_stan_sp5_0h <- stan_occu(~scale(hour)~1, data=umf_y_full_5, chains=3, iter=500, cores=3)
fit_stan_sp5_0c <- stan_occu(~scale(canopy)~1, data=umf_y_full_5, chains=3, iter=500, cores=3)
fit_stan_sp5_1 <- stan_occu(~scale(hour)~scale(elevation), data=umf_y_full_5, chains=3, iter=500, cores=3)
fit_stan_sp5_2 <- stan_occu(~scale(hour)~scale(AGB_Spawn), data=umf_y_full_5, chains=3, iter=500, cores=3)
fit_stan_sp5_3 <- stan_occu(~scale(hour)~scale(roads), data=umf_y_full_5, chains=3, iter=500, cores=3)
fit_stan_sp5_4 <- stan_occu(~scale(hour)~scale(carbon), data=umf_y_full_5, chains=3, iter=500, cores=3)
fit_stan_sp5_5 <- stan_occu(~scale(hour)~scale(NDVI), data=umf_y_full_5, chains=3, iter=500, cores=3)
fit_stan_sp5_6 <- stan_occu(~scale(hour)~scale(canopy), data=umf_y_full_5, chains=3, iter=500, cores=3)


mods5 <- fitList(fit_stan_sp5_0,
                fit_stan_sp5_0h,
                fit_stan_sp5_0c,
                fit_stan_sp5_1,
                fit_stan_sp5_2,
                fit_stan_sp5_3,
                fit_stan_sp5_4,
                fit_stan_sp5_5)

# model selection
round(modSel(mods5), 3)

# plot top model
print("Hour")
plot_effects(fit_stan_sp5_2, "det") # Detection
print("AGB_Spawn")
plot_effects(fit_stan_sp5_2, "state") # Occupancy

```


### Spatial Prediction

```{r}
#| warning: false
#| message: false

#### predict

# srtm_crop_s <- mask(elev, NDVI)#stack(raster(elev))#, 
# scale(roughness)) # scale altitud
# names(srtm_crop_s) <- c("elevation")#, "roughness")
# crs(srtm_crop_s) <- "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0"
nd5 <- mask(AGB_Spawn, NDVI) #data.frame(srtm_crop_s, hour=0.5)
names(nd5) <- "AGB_Spawn"
# newd <- data.frame(srtm_crop_s, hour=0.5)
pred_psi_s_unm <-predict(fit_unm_sp5_2b, type="state", newdata=nd5)
#pred_psi_s_stan <-predict(fit_stan_sp2_1, submodel="state", newdata=nd, na.rm=FALSE)
#pred_psi_ras <- raster(srtm_crop_s) # make raster = to elev 
#pred_psi_ras[] <- pred_psi_s_stan$Predicted # drape on top


# pred_psi_r <- pred_psi_s # * sd(full_covs$elevation) + mean(full_covs$elevation)
# crs(pred_psi_ras) <- "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0"
clr <- colorRampPalette(brewer.pal(9, "YlGn"))
# mapview (pred_psi_r[[1]], col.regions = clr,  legend = TRUE, alpha=0.7)
# plot(pred_psi_s[[1]], main="Occupancy")
levelplot(pred_psi_s_unm[[1]], par.settings = YlOrRdTheme(), margin=FALSE, main="Ocupancy")

pal = mapviewPalette("mapviewSpectralColors")
mapview(raster(pred_psi_s_unm[[1]]), map.types = c("Esri.WorldImagery"), alpha=0.5, col.regions = pal(10)) + mapview(puntos_sf)


```






# *Aramus guarauna*

### One species, single season


```{r}
#| warning: false
#| message: false

# Make UMF object
umf_y_full_1<- unmarkedFrameOccu(y= as.data.frame(parita$occur[[1]])) #Aramus guarauna
siteCovs(umf_y_full_1) <- data.frame(elevation =  covs_raster_val$elevation,
                                   AGB_Spawn = covs_raster_val$AGB_Spawn,
                                   roads = covs_raster_val$roads,
                                   carbon = covs_raster_val$carbon_stock,
                                   NDVI = covs_raster_val$NDVI,
                                   canopy = covs_raster_val$canopy
) # data.frame(Elev=full_covs$Elev) # Full


plot(umf_y_full_1, main="Aramus guarauna")  #""Buteogallus anthracinus no converge

# detection
obsCovs(umf_y_full_1) <- list(hour = matrix(parita$dete[[1]]),
                              canopy = covs_raster_val$canopy ) # hora

# Double formula: first part is for detection, second for occupancy
#  form <- ~detCov1 + detCov2 ~habCov1 + habCov2 + RSR(x, y, threshold=1)

# fit unmarked
fit_unm_sp1_0 <- unmarked::occu(~1~1, data=umf_y_full_1) # Model did not converge!
fit_unm_sp1_0h <- unmarked::occu(~hour~1, data=umf_y_full_1) # Does work but not comparable!
fit_unm_sp1_0c <- unmarked::occu(~scale(canopy)~1, data=umf_y_full_1) # Does work!
fit_unm_sp1_1 <- unmarked::occu(~1~scale(elevation), data=umf_y_full_1)
fit_unm_sp1_2 <- unmarked::occu(~1~scale(AGB_Spawn), data=umf_y_full_1)
fit_unm_sp1_3 <- unmarked::occu(~1~scale(roads), data=umf_y_full_1) # ok!
fit_unm_sp1_4 <- unmarked::occu(~1~scale(carbon), data=umf_y_full_1)
fit_unm_sp1_5 <- unmarked::occu(~1~scale(NDVI), data=umf_y_full_1)
fit_unm_sp1_6 <- unmarked::occu(~1~scale(canopy), data=umf_y_full_1)
fit_unm_sp1_2b <- unmarked::occu(~1~scale(AGB_Spawn), data=umf_y_full_1)


# fit list for detection
fms1<-fitList("p(.) Ocu(.)"=fit_unm_sp1_0,
              #"p(h) Ocu(.)"=fit_unm_sp1_0h, # NA problem!
              "p(c) Ocu(.)"=fit_unm_sp1_0c)#,
#              "p(.) Ocu(elev)"=fit_unm_sp2_1,
#              "p(.) Ocu(slope)"=fit_unm_sp2_2,
#              "p(.) Ocu(aspect)"=fit_unm_sp2_3,
#              "p(.) Ocu(roughness)"=fit_unm_sp2_4,
#              "p(elev^2) Ocu(elev^2)"=fit_unm_sp2_5 )

# model selection detection unmarked
modSel(fms1) #bad!.. null  model ranks better

# fit list for detection
fms2<-fitList("p(.) Ocu(.)"=fit_unm_sp1_0,
              #"p(h) Ocu(.)"=fit_unm_sp2_0h, #,
              "p(c) Ocu(.)"=fit_unm_sp1_0c,
              "p(.) Ocu(elev)"=fit_unm_sp1_1,
              "p(.) Ocu(AGB_Spawn)"=fit_unm_sp1_2,
              "p(.) Ocu(roads)"=fit_unm_sp1_3,
              "p(.) Ocu(carbon)"=fit_unm_sp1_4,
              "p(.) Ocu(NDVI)"=fit_unm_sp1_5,
              "p(.) Ocu(canopy)"=fit_unm_sp1_6)

# model selection detection unmarked
modSel(fms2)


# fit stan
fit_stan_sp1_0 <- stan_occu(~1~1, data=umf_y_full_1, chains=3, iter=500, cores=3)
fit_stan_sp1_0h <- stan_occu(~scale(hour)~1, data=umf_y_full_1, chains=3, iter=500, cores=3)
fit_stan_sp1_0c <- stan_occu(~scale(canopy)~1, data=umf_y_full_1, chains=3, iter=500, cores=3)
fit_stan_sp1_1 <- stan_occu(~scale(hour)~scale(elevation), data=umf_y_full_1, chains=3, iter=500, cores=3)
fit_stan_sp1_2 <- stan_occu(~scale(hour)~scale(AGB_Spawn), data=umf_y_full_1, chains=3, iter=500, cores=3)
fit_stan_sp1_3 <- stan_occu(~scale(hour)~scale(roads), data=umf_y_full_1, chains=3, iter=500, cores=3)
fit_stan_sp1_4 <- stan_occu(~scale(hour)~scale(carbon), data=umf_y_full_1, chains=3, iter=500, cores=3)
fit_stan_sp1_5 <- stan_occu(~scale(hour)~scale(NDVI), data=umf_y_full_1, chains=3, iter=500, cores=3)
fit_stan_sp1_6 <- stan_occu(~scale(hour)~scale(canopy), data=umf_y_full_1, chains=3, iter=500, cores=3)


mods5 <- fitList(fit_stan_sp1_0,
                fit_stan_sp1_0h,
                fit_stan_sp1_0c,
                fit_stan_sp1_1,
                fit_stan_sp1_2,
                fit_stan_sp1_3,
                fit_stan_sp1_4,
                fit_stan_sp1_5)

# model selection
round(modSel(mods5), 3)


# plot top model
print("Hour")
plot_effects(fit_stan_sp1_5, "det") # Detection
print("NDVI")
plot_effects(fit_stan_sp1_5, "state") # Occupancy

```


### Spatial Prediction

```{r}
#| warning: false
#| message: false

#### predict

# srtm_crop_s <- mask(elev, NDVI)#stack(raster(elev))#, 
# scale(roughness)) # scale altitud
# names(srtm_crop_s) <- c("elevation")#, "roughness")
# crs(srtm_crop_s) <- "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0"
nd1 <- NDVI #data.frame(srtm_crop_s, hour=0.5)
names(nd1) <- "NDVI"
# newd <- data.frame(srtm_crop_s, hour=0.5)
pred_psi_s_unm <-predict(fit_unm_sp1_5, type="state", newdata=nd1)
#pred_psi_s_stan <-predict(fit_stan_sp2_1, submodel="state", newdata=nd, na.rm=FALSE)
#pred_psi_ras <- raster(srtm_crop_s) # make raster = to elev 
#pred_psi_ras[] <- pred_psi_s_stan$Predicted # drape on top


# pred_psi_r <- pred_psi_s # * sd(full_covs$elevation) + mean(full_covs$elevation)
# crs(pred_psi_ras) <- "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0"
clr <- colorRampPalette(brewer.pal(9, "YlGn"))
# mapview (pred_psi_r[[1]], col.regions = clr,  legend = TRUE, alpha=0.7)
# plot(pred_psi_s[[1]], main="Occupancy")
levelplot(pred_psi_s_unm[[1]], par.settings = YlOrRdTheme(), margin=FALSE, main="Ocupancy")

pal = mapviewPalette("mapviewSpectralColors")
mapview(raster(pred_psi_s_unm[[1]]), map.types = c("Esri.WorldImagery"), alpha=0.5, col.regions = pal(10)) + mapview(puntos_sf)


```

## Información de la sesión en R.

```{r sesion, results='markup'}
print(sessionInfo(), locale = FALSE)
```


